% DEFINED:   sec:conv-forward, sec:conv-backward, eq:conv-grad
% RESOLVED:  sec:mlp-backprop (backward into Part I)
% FORWARD:   sec:conv-digits (Task 5)

\section{Forward pass, in code}\label{sec:conv-forward}

The \texttt{Conv2D} layer in \texttt{scratchnn} holds \texttt{kernels} of
shape $(C_{\text{out}}, C_{\text{in}}, k, k)$ and a \texttt{bias} of shape
$(C_{\text{out}},)$. Inputs and outputs are flat \texttt{list[float]},
indexed as \texttt{c * H * W + r * W + col} for channel \texttt{c}, row
\texttt{r}, column \texttt{col}. The flat layout keeps the rest of the
library working unchanged: to \texttt{ReLU}, \texttt{Tanh}, and
\texttt{Linear}, the conv output is just a flat vector of length
$C_{\text{out}} \cdot H' \cdot W'$, with the $2$-D shape folded into the
indexing rather than carried as a nested list.

The forward pass is five nested loops, with no NumPy:
\begin{lstlisting}
for c_out in range(C_out):
    kernel = kernels[c_out]
    b = bias[c_out]
    for r in range(out_h):
        for col in range(out_w):
            z = b
            for c_in in range(C_in):
                for i in range(k):
                    for j in range(k):
                        z += kernel[c_in, i, j] * x[c_in, r + i, col + j]
            out[c_out, r, col] = z
\end{lstlisting}
Five loops looks unflattering, and it is exactly as expensive as the math
says: $C_{\text{out}} \cdot C_{\text{in}} \cdot k^2$ multiply-adds per output
spatial position, times $H' \cdot W'$ positions. For the digit network of
\Cref{sec:conv-params} that is $4 \cdot 1 \cdot 9 \cdot 36 = 1296$
multiply-adds per forward pass. The nesting makes the expense visible in the
code, which is the pedagogical point: nothing is hidden inside a tensor
call. The input \texttt{x} is cached for the backward pass, the same pattern
\texttt{Linear} uses in \Cref{sec:mlp-backprop}.

This is the largest layer this book writes in pure Python. The five-loop form
is honest about its cost, and at this scale the cost is affordable.

\section{Backward pass, by careful chain rule}\label{sec:conv-backward}

Let $g_{c_{\text{out}}, r, c} = \partial L / \partial z_{c_{\text{out}}, r,
c}$ be the gradient flowing into the output feature map. Three quantities
follow: gradients with respect to the kernel weights, the bias, and the
input.

\textbf{Kernel gradient.} Each weight $W_{c_{\text{out}}, c_{\text{in}}, i,
j}$ is read at \emph{every} output position $(r, c)$ in the forward pass. By
the chain rule its gradient sums over those positions:
\begin{equation}\label{eq:conv-grad}
  \frac{\partial L}{\partial W_{c_{\text{out}}, c_{\text{in}}, i, j}}
  = \sum_{r, c} g_{c_{\text{out}}, r, c}\,
                x_{c_{\text{in}},\, r+i,\, c+j}.
\end{equation}
This is the \texttt{Linear} formula $g_i\, x_j$ from \Cref{sec:mlp-backprop},
summed over the spatial axes. The reuse of $W$ in the forward pass is exactly
what causes its gradient to accumulate in the backward pass.

\textbf{Bias gradient.} The bias adds to every output position on its feature
map equally, so its gradient is the sum of the incoming gradients there:
\[
  \frac{\partial L}{\partial b_{c_{\text{out}}}}
  = \sum_{r, c} g_{c_{\text{out}}, r, c}.
\]

\textbf{Input gradient.} The pixel $x_{c_{\text{in}}, r', c'}$ was read by the
output positions $(r, c)$ with $r + i = r'$ and $c + j = c'$, that is $(r, c)
= (r' - i, c' - j)$. Summing over all kernels and the valid positions where
the pixel was used,
\[
  \frac{\partial L}{\partial x_{c_{\text{in}}, r', c'}}
  = \sum_{c_{\text{out}}} \sum_{i, j}
    W_{c_{\text{out}}, c_{\text{in}}, i, j}\,
    g_{c_{\text{out}},\, r'-i,\, c'-j},
\]
with out-of-range terms dropped. This is the spatial transpose of the forward
pass; in convolutional-network jargon it is a \emph{transposed convolution}.

All three fall out of one pass. Here is the actual \texttt{Conv2D.backward}
from \texttt{scratchnn}. The offset variables are flat-index bookkeeping (the
$2$-D shape folded into a \texttt{list[float]}, channel-major then row-major);
the three lines that carry the math are the \texttt{+=} accumulations, into
the bias, into the shared kernel, and into the input:
\begin{lstlisting}
def backward(self, g):
    k, in_w, in_hw = self.k, self.in_w, self.in_h * self.in_w
    out_w, out_hw = self.out_w, self.out_h * self.out_w
    dx = [0.0] * len(self.x)
    for c_out in range(self.out_channels):
        kernel, dkernel = self.kernels[c_out], self.dkernels[c_out]
        for r in range(self.out_h):
            for col in range(self.out_w):
                g_rc = g[c_out * out_hw + r * out_w + col]
                self.dbias[c_out] += g_rc                    # dL/db
                for c_in in range(self.in_channels):
                    ox = c_in * in_hw + r * in_w
                    ok = c_in * k * k
                    for i in range(k):
                        for j in range(k):
                            xidx = ox + i * in_w + col + j
                            kidx = ok + i * k + j
                            dkernel[kidx] += g_rc * self.x[xidx]  # dL/dW
                            dx[xidx]      += g_rc * kernel[kidx]   # dL/dx
    return dx
\end{lstlisting}

The observation that matters: every accumulation in \texttt{Conv2D} is the
\emph{same} \texttt{+=} pattern the library already uses for mini-batch
gradients in \texttt{Linear} (\Cref{sec:mlp-backprop}). The mini-batch sums
per-example gradients; the convolution sums per-position gradients. Because
one kernel weight is read at every output position in \texttt{forward}, the
line \texttt{dkernel[kidx] += ...} fires at every position in
\texttt{backward}. Weight sharing and gradient accumulation are one fact read
forward and backward: the same operator, at a finer grain.

The \texttt{Layer} contract from \Cref{sec:mlp-backprop} carries over
unchanged. \texttt{Conv2D} implements \texttt{forward}, \texttt{backward}, and
\texttt{parameters}, and the generic optimizer, written once, never learns
that a convolution exists. A numerical gradient check confirms the hand-derived
backward: the worst relative error is on the order of $10^{-9}$ across small
random inputs, well under the standard $10^{-4}$ tolerance, regenerated by the
chapter's paired notebook.
