% DEFINED:   ch:cnn
% CONTAINS:  section/equation/figure labels live in the \input'd files below
%            (sec:conv-unit, sec:weight-sharing, sec:conv-params,
%            sec:conv-forward, sec:conv-backward, eq:conv-grad,
%            sec:conv-digits, sec:permuted-pixel, fig:permuted-pixel,
%            sec:cnn-bias, sec:pure-python-boundary)
% RESOLVED:  ch:foundations, sec:inductive-biases (backward into Part I)
% CROSS-CHAPTER FORWARD (prose-only, no \Cref): Chapter 4 (sequences /
%            the fixed-context language model), Chapter 6 (the Transformer,
%            where pure Python ends)

\chapter[Convolutional Networks]{Convolutional Networks: Locality and
Translation Equivariance}%
\label{ch:cnn}

\Cref{ch:foundations} named two axes at which a prior enters a supervised
model: the \textbf{output head}, which fixes what the raw output means,
and the \textbf{architecture}, which fixes how the network computes its
features. Part~I followed the output-head axis. Part~II follows the
other one, and this chapter takes its first non-trivial step.

The architecture is a prior over how features compose. A multilayer
perceptron makes the weakest possible such commitment: every input
coordinate may interact with every other on equal footing, and which
interactions matter is left entirely to the data. That generality is a
strength on tabular features with no inherent geometry. On an image it is
a waste. Pixels are arranged in a plane, neighbors mean something,
and a digit is the same digit a few pixels to the left. An MLP knows none
of this; it has to relearn the plane from scratch every time.

A convolutional network is what we build when we refuse to throw that
structure away. It bakes two assumptions into the operator itself rather
than leaving them to be learned. \textbf{Locality}: a unit reads a small
neighborhood, not the whole input. \textbf{Translation equivariance}: the
same unit is applied at every position, so a feature detected in one place
is detected everywhere. The whole chapter is the unpacking of that one
move, in math, in code, and in a falsification experiment. It closes by
marking a boundary: the convolution is the last large layer this book
writes in pure Python.

\input{chapters/ch03/01-conv-and-sharing}
\input{chapters/ch03/02-conv-code}
\input{chapters/ch03/03-digits-and-control}
\input{chapters/ch03/04-bias-and-boundary}
\input{chapters/ch03/bib-notes}
