% DEFINED:   sec:mlp-backprop, eq:linear-grads
% RESOLVED:  sec:linear-limits, sec:nn-as-function
% FORWARD:   sec:gradcheck (Task 8)

\section[The multilayer perceptron]{The multilayer perceptron, and what depth and non-linearity buy}%
\label{sec:mlp-backprop}

The multilayer perceptron (MLP) stacks \texttt{Linear} layers with a
non-linear \textbf{activation} between them (\texttt{Tanh} or
\texttt{ReLU}). Without the non-linearity, stacked linear maps collapse:
two affine maps composed are again affine, and no number of stacked
\texttt{Linear} layers without non-linearities buys anything over one.
The fix from \Cref{sec:linear-limits} is therefore not to add more
linear layers; it is to add a single non-linear layer between them.

\begin{lstlisting}
mlp = Network([Linear(2, 8), Tanh(),
               Linear(8, 1)], loss=SigmoidBCE())
mlp.fit(xor_X, xor_Y, epochs=4000, lr=0.5, batch_size=4)
\end{lstlisting}

Training this 2-8-1 network on XOR (seed~42) reaches a final loss below
$0.001$: the MLP fits XOR exactly. Mechanically, the hidden layer
constructs intermediate features that \emph{are} linearly separable. One
natural decomposition for XOR is
\[
  \mathrm{XOR}(x_1, x_2) = \mathrm{OR}(x_1, x_2) - \mathrm{AND}(x_1, x_2),
\]
so a hidden unit detecting OR and another detecting AND give the output
layer a linearly separable problem. The training run discovers something
equivalent.

The general fact is the \textbf{universal approximation theorem}
(\cite{cybenko1989}; \cite{hornik1989}): a feedforward network with one
hidden layer of sufficient width and any non-polynomial activation can
approximate any continuous function on a compact set to arbitrary
precision. \emph{Depth is not required for expressivity.} One hidden
layer suffices. What deeper networks buy is \emph{efficient}
representation: many natural functions that one wide hidden layer can
approximate at exponential cost are representable at polynomial cost
with a few stacked hidden layers. That is the empirical case for depth;
the theoretical case for hidden layers at all is XOR.

\subsection*{Backpropagation as the local chain rule}

The training rule for an MLP is \textbf{backpropagation}. The trick is
to make the chain rule \emph{local}: each layer implements
\texttt{forward} (input to output) and \texttt{backward} (gradient with
respect to the output, to gradient with respect to the input), plus
\texttt{parameters}, which exposes its weights and their gradients. A
layer never sees the rest of the network. That contract is the entire
base class:

\begin{lstlisting}
class Layer:
    def forward(self, x):
        raise NotImplementedError

    def backward(self, g):
        """Given g = dL/d(output), return dL/d(input).
        Layers with parameters also accumulate their parameter grads."""
        raise NotImplementedError

    def parameters(self):
        """A list of (values, grads) pairs: each a parameter vector
        and its same-length gradient vector."""
        return []
\end{lstlisting}

\texttt{Linear} ($\vy = W\vx + \mathbf{b}$) is the only layer with
parameters. Its \texttt{backward} writes out three gradients:
\begin{equation}\label{eq:linear-grads}
  \frac{\partial L}{\partial W_{ij}} = g_i x_j,
  \qquad
  \frac{\partial L}{\partial b_i} = g_i,
  \qquad
  \frac{\partial L}{\partial x_j} = \sum_i W_{ij}\, g_i.
\end{equation}
That is exactly what the code says. Note the \texttt{+=}: parameter
gradients \emph{accumulate}, so a mini-batch can sum per-example
contributions before the optimizer steps.

\begin{lstlisting}
class Linear(Layer):
    def forward(self, x):
        self.x = x
        return [dot(w, x) + b for w, b in zip(self.weights, self.bias)]

    def backward(self, g):
        for i, gi in enumerate(g):
            for j, xj in enumerate(self.x):
                self.dweights[i][j] += gi * xj     # dL/dW_ij = g_i x_j
            self.dbias[i] += gi                    # dL/db_i  = g_i
        return [sum(self.weights[i][j] * g[i] for i in range(len(g)))
                for j in range(len(self.x))]       # dL/dx_j = sum_i W_ij g_i
\end{lstlisting}

There is no matrix type. A vector is a \texttt{list[float]}, \texttt{dot}
is the only named helper, and the outer product and transposed mat-vec
above are written inline as comprehensions. The two activations are
smaller still, each matching the derivatives one-to-one ($1 - \tanh^2$
for \texttt{Tanh}, the positive-part mask for \texttt{ReLU}):

\begin{lstlisting}
class Tanh(Layer):
    def forward(self, x):
        self.out = [math.tanh(xi) for xi in x]
        return self.out
    def backward(self, g):
        return [gi * (1.0 - oi * oi) for gi, oi in zip(g, self.out)]

class ReLU(Layer):
    def forward(self, x):
        self.positive = [xi > 0.0 for xi in x]
        return [xi if xi > 0.0 else 0.0 for xi in x]
    def backward(self, g):
        return [gi if p else 0.0 for gi, p in zip(g, self.positive)]
\end{lstlisting}

The \texttt{Network} threads these together. \texttt{forward} runs the
layers in order and caches the logits; \texttt{backward} seeds the
gradient from the loss and pushes it back through the layers in reverse:

\begin{lstlisting}
class Network:
    def forward(self, x):
        for layer in self.layers:
            x = layer.forward(x)
        self.logits = x
        return x

    def backward(self, y):
        g = self.loss.grad(self.logits, y)    # the familiar p - y
        for layer in reversed(self.layers):
            g = layer.backward(g)
        return g
\end{lstlisting}

The optimizer is written \emph{once}, generically, on top of
\texttt{parameters()}. There is no per-layer \texttt{step} and no
\texttt{Optimizer} class. \texttt{zero\_grad} clears every gradient;
\texttt{step} applies the \emph{mean} gradient over a batch of $n$ by
dividing by $n$, which is the counterpart to the accumulating
\texttt{+=} in \texttt{backward}:

\begin{lstlisting}
    def zero_grad(self):
        for _values, grads in self.parameters():
            for k in range(len(grads)):
                grads[k] = 0.0

    def step(self, lr, n):
        for values, grads in self.parameters():
            for k in range(len(values)):
                values[k] -= (lr / n) * grads[k]
\end{lstlisting}

The key realization: the MLP introduced \textbf{no new math} beyond what
a single linear unit needed. The chain rule applied locally, composed
across layers, gives gradients for any depth. Depth is just more
composition of the same local steps, and the optimizer never has to know
how many layers there are. \Cref{sec:gradcheck} verifies all this
numerically.
