% DEFINED:   sec:logistic, sec:softmax, eq:py-binary, eq:py-softmax
% RESOLVED:  sec:linear-limits, sec:mlp-backprop, sec:nn-as-function
% FORWARD:   sec:inductive-biases (Task 7)

\section{Logistic regression}\label{sec:logistic}

A logistic-regression model scores an input $\vx$ with a weighted sum,
the \textbf{logit}:
\[
  z = \vw \cdot \vx + b.
\]
It squashes that score into a probability with the \textbf{sigmoid}:
\[
  p = \sigma(z) = \frac{1}{1 + e^{-z}}.
\]
Here $p$ is the model's estimate of $P(y = 1 \mid \vx)$. The loss that
penalizes disagreement between $p$ and a binary label $y \in \{0, 1\}$
is \textbf{binary cross-entropy} (BCE):
\[
  L = -\bigl[y \log p + (1-y) \log(1-p)\bigr].
\]
Differentiating with respect to the logit $z$ gives a clean result:
\begin{equation}\label{eq:py-binary}
  \frac{\partial L}{\partial z} = p - y.
\end{equation}
Prediction minus target. Every weight gradient follows by the chain
rule, $\partial L / \partial w_j = (p - y)\, x_j$, and we descend:
$\vw \leftarrow \vw - \eta\, \partial L / \partial \vw$.

In code, logistic regression is one \texttt{Linear} layer with a
\texttt{SigmoidBCE} loss:
\begin{lstlisting}
net = Network([Linear(n_features, 1)], loss=SigmoidBCE())
\end{lstlisting}
The loss owns the output activation. The \texttt{Network} emits the raw
logit; \texttt{SigmoidBCE} turns it into a probability and a gradient.
Its \texttt{grad} is the $p - y$ of \Cref{eq:py-binary}; its
\texttt{value} computes binary cross-entropy straight from the logit in a
numerically stable form (discussed in \Cref{sec:stability}), never by
first forming $p$ and taking a log:

\begin{lstlisting}
class SigmoidBCE(Loss):
    def value(self, logits, y):
        z = logits[0]
        # max(z,0) - z*y + log1p(exp(-|z|))
        return max(z, 0.0) - z * y + math.log1p(math.exp(-abs(z)))

    def grad(self, logits, y):
        return [sigmoid(logits[0]) - y]      # p - y

    def probs(self, logits):
        return [sigmoid(logits[0])]
\end{lstlisting}

This is structurally the same SLP from \Cref{sec:linear-limits}, with a
probability interpretation. The SLP's limits still apply: logistic
regression can solve any linearly separable binary classification but
cannot solve XOR. To solve XOR with sigmoid output, swap the body for
the MLP from \Cref{sec:mlp-backprop}:
\begin{lstlisting}
net = Network([Linear(2, 8), Tanh(),
               Linear(8, 1)], loss=SigmoidBCE())
\end{lstlisting}
The output head (sigmoid + BCE) and the architecture (linear vs.~MLP)
are independent choices, a theme that becomes the load-bearing principle
of \Cref{sec:inductive-biases}.

\section{Multi-class regression via softmax}\label{sec:softmax}

With $K$ classes the model produces $K$ logits, one weighted sum per
class. The logit vector $\vz$ becomes a probability distribution through
the \textbf{softmax}:
\[
  p_i = \frac{e^{z_i}}{\sum_j e^{z_j}}.
\]
The matching loss is \textbf{categorical cross-entropy}, $L = -\log p_y$,
where $y \in \{0, \ldots, K-1\}$ is the index of the correct class. Its
gradient with respect to the logits is, again:
\begin{equation}\label{eq:py-softmax}
  \frac{\partial L}{\partial z_i} = p_i - \mathbf{1}[i = y].
\end{equation}
That is $\vp - \vy$ with $\vy$ read as a one-hot vector. The same
expression as the binary case (\Cref{eq:py-binary}). That is no
coincidence: binary logistic regression \emph{is} two-class softmax with
one logit pinned to zero (softmax depends only on logit differences, so
one logit may be fixed freely). The binary sigmoid is the $K = 2$
softmax in disguise.

In code, softmax regression is one \texttt{Linear} layer with a
categorical head:
\begin{lstlisting}
net = Network([Linear(n_features, K)], loss=SoftmaxCrossEntropy())
\end{lstlisting}
The head mirrors \texttt{SigmoidBCE} exactly, one level up. \texttt{value}
uses the stable $\mathrm{logsumexp}(\vz) - z_y$ form of
$-\log \mathrm{softmax}(\vz)_y$; \texttt{grad} is $\vp$ with $1$
subtracted at the true class:

\begin{lstlisting}
class SoftmaxCrossEntropy(Loss):
    def value(self, logits, y):
        # Stable: logsumexp(z) - z[y]
        return logsumexp(logits) - logits[y]

    def grad(self, logits, y):
        p = softmax(logits)
        p[y] -= 1.0
        return p

    def probs(self, logits):
        return softmax(logits)
\end{lstlisting}

On the 8$\times$8 UCI optical-recognition digits
(1\,437 training, 360 test, 10 classes, stratified 80/20 split,
seed~42), this one-layer softmax model reaches \textbf{94.4\%} test
accuracy after 30 epochs of mini-batch SGD. Linear classifiers on this
dataset are unexpectedly competitive, which is itself a hint about the
data: the $4 \times 4$ block-counting that produced these 8$\times$8
features has already done most of the work.

Swap the body for an MLP:
\begin{lstlisting}
net = Network([Linear(64, 32), Tanh(),
               Linear(32, 10)], loss=SoftmaxCrossEntropy())
\end{lstlisting}
This 64-32-10 MLP reaches \textbf{97.2\%} test accuracy after 50 epochs.
The MLP fits the training set to \textbf{99.3\%} and generalizes
somewhat worse; the gain over the linear classifier is real but modest
(2.8 percentage points). Why is the gap so small, given that the MLP is
so much more expressive than logistic regression? The answer is in
\Cref{sec:inductive-biases}.
