% DEFINED:   sec:logits-logprobs, eq:logodds
% RESOLVED:  sec:softmax
% FORWARD:   sec:stability (Task 8)

\section{Logits are log-probabilities, up to a constant}\label{sec:logits-logprobs}

Take the log of the softmax from \Cref{sec:softmax}:
\[
  \log p_i = z_i - \log \sum_j e^{z_j}.
\]
The logits are \emph{not} the log-probabilities. They are the
log-probabilities plus a shared additive term $\log \sum_j e^{z_j}$,
the same for every class $i$. The logits are \textbf{unnormalized
log-probabilities}.

That constant cannot be recovered from $\vp$. Softmax is unchanged by
adding any constant $c$ to every logit: $e^{z_i + c}$ has a factor
$e^c$ on top and bottom, and it cancels. Training therefore can never
pin the constant down. What training \emph{does} pin down is logit
\textbf{differences}:
\begin{equation}\label{eq:logodds}
  z_i - z_j = \log p_i - \log p_j = \log \frac{p_i}{p_j},
\end{equation}
the log-odds between classes. The constant is \textbf{gauge}: free,
unobservable, with no effect on the predicted distribution.

One sharp way to see that logits are not raw log-probabilities: a
probability is at most $1$, so a log-probability is at most $0$. But
nothing stops a network's logits from being positive. Put a ReLU just
before the output, and every logit is at least $0$. They still describe
a perfectly good distribution, because the additive constant absorbs
the offset. Logits live on a different scale than log-probabilities;
only their differences are meaningful.

This gauge freedom returns in \Cref{sec:stability} as the key to
numerical stability: since softmax does not care about a shared shift,
we can spend that freedom on keeping the exponentials in a safe range.
