% DEFINED:   sec:inductive-biases
% RESOLVED:  sec:softmax
% FORWARD REFS (prose-only, no \Cref): Chapter 2 on output heads,
%   Part II on architecture (CNN, fixed-context LM, RNN, Transformer),
%   Chapter 8 on reinforcement learning

\section{Inductive biases}\label{sec:inductive-biases}

The MLP on digits is essentially saturated at 97.2\% test accuracy
(seed~42, 80/20 split), even though it is plenty expressive: it fits the
training set to 99.3\%. A model that fits the training data almost
perfectly yet plateaus below it on held-out data is not
capacity-limited. The bottleneck is the \textbf{prior} the architecture
and the loss commit to.

Inductive bias enters at two distinct points in every supervised model:

\begin{enumerate}
\item \textbf{The output head}: a choice of link function and matching
  loss. Identity plus MSE assumes Gaussian targets; sigmoid plus BCE
  assumes Bernoulli; softmax plus cross-entropy assumes Categorical. The
  link function is a prior about the data-generating process on the
  output side. Chapter~2 takes this axis up in depth.
\item \textbf{The architecture}: a choice of how the network computes
  features. An MLP assumes every input feature is independent of every
  other and learns their relationships from data. A convolutional network
  assumes nearby pixels matter together and the same detector should fire
  anywhere. A recurrent network assumes the same computation runs at every
  step of a sequence. A Transformer assumes positions interact pairwise
  through attention. Part~II of this book walks through each of these
  priors.
\end{enumerate}

Both axes are inductive biases. Both improve sample efficiency when
matched to the data and hurt when mismatched. The two are independent:
any architecture can pair with any output head, as long as the head's
input dimension matches the architecture's output.

For the 64 pixels of a digit, the MLP architecture says every feature
is independent of every other. That is a strong claim, and it happens to
be wrong. Pixel $(3, 4)$ and pixel $(3, 5)$ are neighbors; the MLP
treats them as completely unrelated and has to relearn their
relationship from 1\,437 examples. Pixel $(3, 4)$ and pixel $(7, 0)$
are unrelated, but the MLP cannot tell that apart from $(3, 4)$ and
$(3, 5)$ either.

Different architectures commit to different invariances:

\begin{itemize}
\item A \textbf{convolutional network} (CNN) commits to \textbf{locality}
  and \textbf{translation equivariance}: nearby pixels matter together,
  and a feature detector that fires at one position should fire at any
  position. The right prior for image data.
\item An \textbf{RNN} commits to \textbf{time-translation equivariance}:
  the same computation runs at every step of a sequence, and the hidden
  state summarizes the past. The right prior for sequential data.
\item A \textbf{Transformer} commits to \textbf{permutation equivariance}
  over positions, with positional encoding added back to recover a
  sequence prior. Pairwise interactions are explicit through attention
  rather than recurrence.
\end{itemize}

These are bets. A CNN beats an MLP on images \emph{because} its prior
matches the data. On tabular data with no spatial structure, a CNN loses
to an MLP, because the spatial prior is wrong. The right prior is
empirical, not universal.

There is a data-side counterpart. \textbf{Data augmentation} expresses
the same prior from the data side instead of the architecture side.
Random rotations of training images approximate the prior ``the answer
should not depend on rotation'' through the loss: a rotation-equivariant
architecture and rotation-augmented data are two routes to the same
prior, and they often compose.

A concrete illustration for digits: a single $3 \times 3$ convolutional
filter applied across the $8 \times 8$ image with 16 output channels has
\textbf{160 parameters}. The MLP we trained has \textbf{2\,410
parameters}. The CNN ends up beating the MLP with one-fifteenth the
parameters, \emph{because} it has committed to the right prior and does
not have to learn translation invariance from scratch for every position.
That is the whole story of inductive bias in one number. (The exact
counts are computed in the paired notebook, not asserted.)

When you do not know the structure, an MLP is the safe bet on the
architecture side and identity plus MSE is the safe bet on the output
side: both commit to nothing, and pay for that with data efficiency.

The next chapter takes up the output-head axis in full: how each output
head corresponds to a distributional assumption about the target, why the
$p - y$ gradient recurs, and how the two axes compose. The architecture
axis runs through Part~II. A closing chapter on reinforcement learning
asks how much of this inductive-bias frame survives the shift from
supervised learning to learning from scalar reward.
