% DEFINED:   (none; unnumbered section)
% RESOLVED:  \cite keys only
% FORWARD:   (none)

\section*{Bibliographic Notes}

The algebraic impossibility argument for XOR is from
\textcite{minsky1969perceptrons}. Their book demonstrated that
single-layer perceptrons cannot represent parity or connected figures,
which redirected neural network research toward multi-layer
architectures for the following decade.

Universal approximation for shallow networks with sigmoidal activations
was proved independently by \textcite{cybenko1989} and
\textcite{hornik1989}. Both results apply to a single hidden layer of
sufficient width; they say nothing about the efficiency of depth. The
empirical case for depth over width is a separate (and still active)
story.

Backpropagation in the form the machine-learning community uses it was
presented by \textcite{rumelhart1986backprop}, though the underlying
chain-rule recursion was known earlier in control theory and
numerical analysis.

The UCI optical-recognition dataset (\cite{uci1998digits}) is the
8$\times$8 image version derived from the handwritten-digits recordings
at the National Institute of Standards and Technology. It ships with
\texttt{sklearn.datasets.load\_digits} as a 1\,797-sample subset of the
full UCI release.

For the full story of automatic differentiation as a discipline distinct
from symbolic and numerical differentiation, the survey by
\textcite{baydin2018autodiff} is the standard reference. It places
reverse-mode AD (which is what backpropagation is) in the context of
the broader field and covers both the forward and reverse accumulation
modes with worked examples.

\textcite{goodfellow2016dlbook} covers the material of this chapter at
book length, with particular depth on regularization, optimization, and
the practical engineering of deep networks.
