% DEFINED:   sec:rnn-prior, sec:rnn-cell
% RESOLVED:  ch:foundations, ch:cnn, ch:fixed-context-lm (backward into
%            Part I and earlier Part II)
% FORWARD:   (none; the vanishing-gradient limit is a later section, prose-only)

\section{The prior}\label{sec:rnn-prior}

A sequence is an ordered series of inputs $x_1, x_2, \ldots, x_T$. The
contents might be characters in a string, tokens in a sentence, ticks of a
time series, frames of a video, or pixels of an image read out in raster
order. The structure that makes it a sequence is that time matters: $x_t$
sits closer to $x_{t+1}$ than to $x_{t+5}$, and a computation that is useful
at one step should be useful at any step.

Two assumptions turn that structure into a prior, and they have the same
shape as the convolutional network's. The first is locality in time: the
prediction at step $t$ leans most heavily on the recent past, not on every
token from the start of the sequence with equal weight. The second is
time-translation equivariance: shift the input sequence by $\Delta t$ and
the output shifts by the same $\Delta t$. The same computation applies at
every step, so the model's weights do not depend on the absolute timestamp.
Neither assumption is free; each is a bet that the data respects a symmetry,
and the next section makes the bet concrete.

A multilayer perceptron applied to a flattened sequence has neither prior.
It treats $x_t$ at every $t$ as an independent feature, with its own slab of
weights for each position. Given enough data it can still learn the right
function, but it learns each position from scratch and does not carry what
it learned at one offset over to another. A pattern seen only early in
training at one position is a pattern it has never seen at any other.

There are two ways to wire the temporal priors into an architecture, and
they part on a single question: how much of the past does a prediction get
to see. \Cref{ch:fixed-context-lm} took the first answer. Look at a fixed
window of the last $N$ tokens, share the embedding across positions, and
feed a small head. That is a bounded window, and everything older than $N$
is gone by construction. The recurrent network takes the second answer.
Rather than a fixed slice of the past, it carries a hidden state vector
$h_t$ that summarizes the past, and it updates that summary one step at a
time. The next state $h_{t+1}$ is computed from the current state $h_t$ and
the current input $x_{t+1}$, using the same weights at every step. Because
the state is a running summary rather than a window, the model can in
principle let a prediction depend on an input from arbitrarily far back. In
practice that reach has a hard limit, which a later section in this chapter
spends its whole length on.

\section{The cell}\label{sec:rnn-cell}

A vanilla recurrent cell is one equation:
\[
  h_t = \tanh\!\bigl(W_{xh}\, x_t + W_{hh}\, h_{t-1} + b_h\bigr),
\]
where $W_{xh}$ is the input-to-hidden weight matrix, $W_{hh}$ is the
recurrent hidden-to-hidden matrix, $b_h$ is the bias, and $\tanh$ is applied
componentwise to the pre-activation vector. The cell's output is $h_t$
itself. A separate \texttt{Linear} layer projects $h_t$ to logits, or to
whatever shape the task downstream wants, the same division of labor every
chapter has kept: the body produces a representation, and a head interprets
it.

The property that makes this a recurrent network and not just a stack of
distinct layers is that $W_{xh}$, $W_{hh}$, and $b_h$ do not depend on $t$.
One set of weights computes the state update at every step. This is weight
sharing along the time axis, the exact analogue of the convolutional
network's weight sharing across spatial positions in \Cref{ch:cnn}. There
the same kernel slid across an image; here the same update slides along a
sequence.

In code the cell extends the \texttt{Layer} contract from
\Cref{ch:foundations}. A stateless layer (\texttt{Linear}, \texttt{Tanh},
\texttt{ReLU}) has \texttt{forward(x)} returning an output and
\texttt{backward(g)} returning the gradient on its input. A stateful layer
extends both signatures: \texttt{forward(x, state)} takes the previous state
and returns \texttt{(output, new\_state)}, and \texttt{backward(g\_out,
dstate\_next)} takes a gradient on the new state and returns
\texttt{(dL/dx, dL/dstate\_prev)}. For a vanilla cell the new state equals
the output, both are $h_t$, so the forward returns the same vector twice, as
\texttt{(h\_new, h\_new)}. The two protocols share \texttt{parameters()},
so the generic \texttt{step()} and \texttt{zero\_grad()} from
\Cref{ch:foundations} need no change. This is the same split between a
stateful module and a stateless function that production frameworks
navigate.

Parameter storage matches \texttt{Linear} exactly. The cell holds $W_{xh}$
and $W_{hh}$ as lists of rows, one row per hidden unit, alongside the bias
vector, so \texttt{parameters()} yields one (values, gradients) pair per row
and the optimizer steps them without knowing a recurrence exists. Here is
the real forward, with the constructor summarized in a comment:
\begin{lstlisting}
class RNNCell(Layer):
    def __init__(self, input_size, hidden_size):
        # W_xh, W_hh, b_h and their gradient accumulators. Xavier-style
        # init. Stored as list[list[float]], one row per hidden unit, like
        # Linear, so parameters() yields one (values, grads) pair per row.
        # A LIFO cache holds (x, h_prev, h_new) per forward, popped in
        # backward.
        ...

    def forward(self, x, state=None):
        H = self.hidden_size
        if state is None:
            state = [0.0] * H
        h_new = [0.0] * H
        for i in range(H):                       # one row per hidden unit
            z = self.b_h[i]
            for j in range(self.input_size):
                z += self.W_xh[i][j] * x[j]      # W_xh @ x
            for j in range(H):
                z += self.W_hh[i][j] * state[j]  # W_hh @ h_prev
            h_new[i] = math.tanh(z)
        self.cache.append((x, state, h_new))     # LIFO, popped in backward
        return h_new, h_new

    def backward(self, dh_out, dstate_next=None):
        ...  # the BPTT step, shown in full in a later section
        return dx, dh_prev
\end{lstlisting}
The forward is two nested loops, the same matrix-vector shape as
\texttt{Linear}, with the second loop adding the recurrent contribution
$W_{hh}\, h_{t-1}$ on top of the input contribution $W_{xh}\, x_t$. The only
genuinely new line is the cache push. The cell saves the inputs it will need
to differentiate, because the backward pass that undoes this step runs much
later, after the whole sequence has been read. That deferred, accumulated
backward is the subject of the next section.
