% DEFINED:   ch:rnn
% CONTAINS:  section/figure/equation labels live in the \input'd files below
%            (sec:rnn-prior, sec:rnn-cell, sec:time-equivariance, sec:bptt,
%            sec:vanishing-gradients, eq:rnn-jacobian, sec:rnn-alice,
%            fig:ch05-rnn-loss, sec:rnn-pattern, sec:rnn-handoff)
% RESOLVED:  ch:foundations, sec:mlp-backprop, ch:cnn, sec:permuted-pixel,
%            ch:fixed-context-lm (all backward into Part I and earlier Part II)
% CROSS-CHAPTER FORWARD (prose-only, no \Cref): Chapter 6 (the Transformer,
%            "the next chapter"). Not drafted yet; stays prose.

\chapter[Recurrent Networks]{Recurrent Networks: Time-Translation
  Equivariance}%
\label{ch:rnn}

A recurrent network is a multilayer perceptron with a single prior wired
into its weights: the same computation runs at every step of a sequence,
and a hidden state carries a summary of the past forward from one step to
the next. The first move is the convolutional network's move from
\Cref{ch:cnn}, weight sharing across positions, with one substitution.
``Position'' is now an index in time rather than in space.

\Cref{ch:cnn} shared one kernel across spatial positions and got
translation equivariance for images. This chapter makes the same move for
sequences. The math, the implementation, and even the \texttt{+=}
accumulation pattern in the backward pass all transfer directly. Two things
change. The axis the sharing runs along is time instead of space, and the
recurrence adds a wrinkle the spatial version did not have: the layer's
output at one step is also its input at the next, so the chain of
computations forms one long sequential dependency. That dependency is what
lets the state reach back in principle to any earlier input, and it is also
what makes a recurrent network hard to train at long range. By the end of
the chapter we will have earned the right to name vanishing gradients as the
price of the prior.

The plan is the usual one for Part~II. State the prior, build the smallest
model that isolates it, read its real code, derive the gradient by hand,
train it on a corpus small enough to watch, and locate it on the map of
architectural priors. The corpus is the same char-level \emph{Alice}
\Cref{ch:fixed-context-lm} used, and the chapter closes by setting the
recurrent network's unbounded-in-principle memory against that chapter's
bounded window, the head-to-head comparison the fixed-context chapter
deferred to here. Everything is pure-Python \texttt{scratchnn}: a stateful
\texttt{RNNCell}, a \texttt{Linear} head, and \texttt{SoftmaxCrossEntropy},
trained by a hand-derived backpropagation through time.

\input{chapters/ch05/01-prior-and-cell}
\input{chapters/ch05/02-equivariance-bptt-vanishing}
\input{chapters/ch05/03-experiment-and-comparison}
\input{chapters/ch05/04-pattern-and-handoff}
\input{chapters/ch05/bib-notes}
