% DEFINED:   sec:layer2-deref, sec:layer1-aggregator, sec:ablations
% RESOLVED:  (none new)
% FORWARD:   fig:attn-contrast (defined in 04-scaling-the-lens)

\section{Layer 2 attends to the addressed position}\label{sec:layer2-deref}

Start with the dereference, the second half of the hypothesis, because it is the
easiest to see. Take four representative test examples with addresses
$a \in \{0, 4, 5, 7\}$ and extract the layer-2 attention pattern. The row that
matters is row 10, the query at the last input position. Its argmax, the memory
position receiving the most attention, lands exactly on the address every time:

\begin{center}
\begin{tabular}{@{}rrr@{}}
\toprule
Address $a$ & Layer-2 argmax position & Weight on that position \\
\midrule
0 & 0 & 0.698 \\
4 & 4 & 0.973 \\
5 & 5 & 0.881 \\
7 & 7 & 0.851 \\
\bottomrule
\end{tabular}
\end{center}

Every example sends most of position 10's layer-2 attention to the memory
position whose index equals the address. The left panel of
\Cref{fig:attn-contrast} shows one such map: for $a = 4$ the attention mass at
the last query row concentrates almost entirely on column 4, the cell $m_4$.

Four hand-picked examples could be a coincidence, so aggregate. The probe runs
the layer-2 attention for many examples, takes the last row of each, and groups
those rows by the decoded address:

\begin{lstlisting}
def aggregate_last_row(model, X, M=8, A=3):
    """Layer-2 attention from the last position, grouped by address."""
    T = M + A
    by_addr = {a: [] for a in range(2 ** A)}
    for ids in X:
        ctx = ids[:T]
        a = address_value(ctx, M=M, A=A)
        attn_L2 = model.attention_at(ctx, layer=1, head=0)
        last_row = attn_L2[T - 1]
        by_addr[a].append(last_row)
    means = {a: np.mean(np.stack(v), axis=0) for a, v in by_addr.items() if v}
    return means
\end{lstlisting}

Averaged over 500 test examples, partitioned by encoded address, the picture is
clean for every address, not just the four:

\begin{center}
\begin{tabular}{@{}rrr@{}}
\toprule
Address $a$ & Avg weight on $m_a$ & Avg weight on other memory \\
\midrule
0 & 0.795 & 0.205 \\
1 & 0.706 & 0.294 \\
2 & 0.661 & 0.339 \\
3 & 0.708 & 0.252 \\
4 & 0.812 & 0.188 \\
5 & 0.710 & 0.286 \\
6 & 0.729 & 0.269 \\
7 & 0.676 & 0.315 \\
\bottomrule
\end{tabular}
\end{center}

For every address from 0 to 7, the average weight on the correct memory cell
$m_a$ is between 0.66 and 0.81, and the combined weight on all other memory cells
is only 0.19 to 0.34. The mapping from address to attended position is clean and
consistent across the test distribution. Layer 2 is the dereference, exactly as
predicted.

\section{Layer 1 is the address aggregator}\label{sec:layer1-aggregator}

Now the other half: layer 1 should be gathering the address bits into position
10 so that layer 2's query can use them. For the same four examples, here is
where layer 1 at position 10 sends its attention. The table gives the top three
attended positions with weights, and the total mass on the address positions
$\{8, 9, 10\}$ against the memory positions $\{0, \ldots, 7\}$:

\begin{center}
\begin{tabular}{@{}rlrr@{}}
\toprule
Address $a$ & Top-3 (position, weight) & Total on addr & Total on memory \\
\midrule
0 & (8, 0.49), (9, 0.41), (10, 0.05) & 0.956 & 0.044 \\
4 & (8, 0.71), (9, 0.20), (7, 0.06) & 0.935 & 0.065 \\
5 & (8, 0.73), (9, 0.19), (10, 0.06) & 0.986 & 0.014 \\
7 & (8, 0.51), (9, 0.41), (10, 0.04) & 0.965 & 0.035 \\
\bottomrule
\end{tabular}
\end{center}

The pattern is unambiguous. Layer 1 at position 10 puts 93 to 99 percent of its
attention mass on the three address positions. The memory bits, which dominate
the information content of the input by a factor of eight to three, receive
almost none.

There is a subtler thing in the table worth pausing on. Within the address
positions the weights are not uniform. Position 8, the most significant bit
$a_0$, gets the most weight; position 9, $a_1$, less; position 10, the least
significant bit $a_2$, the least. This is sensible rather than incidental. The
most significant bit partitions memory into upper and lower halves, the next bit
selects quadrants, and the least significant bit selects within pairs. A model
that has to write the whole address into a single residual vector through a
linear value map will weight its inputs by how much they matter for the
downstream lookup, and the most significant bit matters most for telling apart
memory positions that are far apart. The hypothesis is vindicated on this side
too: layer 1 is gathering the address, by exactly the
attention-as-content-addressable-lookup mechanism of \Cref{ch:transformer}, with
the position-10 query learning to match the position embeddings of 8 and 9.

\section{Causal confirmation: ablations}\label{sec:ablations}

Everything so far is correlational. We have seen that layer 1 attends to the
address positions and layer 2 attends to the correct memory cell, which is
consistent with the circuit hypothesis. It does not by itself rule out a duller
story: that the model leans mostly on position 10's own embedding plus the bias
structure of the data, with attention going along for the ride. To separate the
two we need a causal test.

The cleanest one is to replace a layer's attention pattern with the
uniform-on-causal-prefix distribution, in which each query position attends with
equal weight to every position at or before it, and then rerun the rest of the
forward pass. This wipes out the layer's content-addressable behavior while
leaving its value and feed-forward machinery untouched. If a layer is doing real
work, the substitution should hurt. The same routine also supports a second
intervention: shuffle the three address bits before the forward pass, which
leaves the bit \emph{set} unchanged but destroys which bit is which.

\begin{lstlisting}
def ablation_shuffle_address(model, X, Y, M=8, A=3, seed=0):
    """Permute the address bits across each example before forward."""
    rng = np.random.default_rng(seed)
    correct = 0
    T = M + A
    for ids, y in zip(X, Y):
        ctx = ids[:T].copy()
        addr = ctx[M:M + A].copy()
        rng.shuffle(addr)
        ctx[M:M + A] = addr
        logits = model.forward(ctx)
        pred = int(np.argmax(logits[-1]))
        if pred == int(y):
            correct += 1
    return correct / len(X)
\end{lstlisting}

The uniform-prefix substitution is built the same way one would expect, by
replacing the cached pattern with a normalized lower-triangular matrix before
the value-weighted sum:

\begin{lstlisting}
uniform = np.tril(np.ones((T, T), dtype=np.float32))
uniform = uniform / uniform.sum(axis=-1, keepdims=True)
\end{lstlisting}

On 500 test examples:

\begin{center}
\begin{tabular}{@{}lr@{}}
\toprule
Intervention & Test accuracy \\
\midrule
Baseline (no intervention) & 1.000 \\
Layer 1 attention $\to$ uniform & 0.658 \\
Layer 2 attention $\to$ uniform & 0.636 \\
Shuffle the three address bits before forward & 0.776 \\
\bottomrule
\end{tabular}
\end{center}

Both layer ablations drop accuracy by a third. Either layer is necessary,
neither alone is sufficient, and both have to be doing the hypothesized job.
That is the causal confirmation the correlational evidence could not give on its
own.

The shuffle ablation is the most instructive of the three. Permuting the address
bits keeps the same multiset of bits but destroys the positional encoding of
which bit is which. If the model depended only on whether each address bit was 0
or 1, a permutation-invariant function of the bits, the shuffle would not hurt.
The drop from 1.000 to 0.776 says the model is using the bits with their
positional roles, exactly as a circuit computing $a = 4 a_0 + 2 a_1 + a_2$ must.
And the fact that accuracy lands at 0.776 rather than at chance is itself
informative. The memory bits are still present, so when many of them happen to
agree the prediction is right regardless of which cell is addressed. The residual
0.776 is the "guess the majority bit" baseline given a scrambled address, and the
0.224 gap below the baseline is the attributable contribution of correctly routed
addressing.
