% DEFINED:   ch:interpretability
% CONTAINS:  section/figure labels live in the \input'd files below
%            (sec:interp-setup, sec:circuit-restated, sec:layer2-deref,
%            sec:layer1-aggregator, sec:ablations, sec:induction-head,
%            sec:icl, sec:small-model-limits, sec:scaling-the-lens,
%            fig:attn-contrast, fig:lookup-by-address, fig:causal-trace,
%            sec:interp-closing)
% RESOLVED:  ch:transformer, sec:one-layer-limit, sec:pe-scale-fix,
%            sec:parallel-axes, sec:three-axes (all backward into Parts I-II)
% CROSS-CHAPTER FORWARD (prose-only, no \Cref): Chapter 8 (reinforcement
%            learning; the closing chapter). Not drafted yet; stays prose.

\chapter[Reverse-Engineering the Pointer Transformer]{Reverse-Engineering
the Pointer Transformer}%
\label{ch:interpretability}

\Cref{ch:transformer} built a two-layer, one-head transformer that solves
the pointer-dereferencing task at $M = 8$ to 100 percent test accuracy. We
argued from the architecture (\Cref{sec:one-layer-limit}) that the model
\emph{must} be composing two attention lookups: the first layer aggregates
the address bits at the last input position, and the second layer uses the
assembled address to attend to the addressed memory bit. That argument was a
prediction. This chapter checks it against the trained weights.

That is the whole move of Part~III. The first two parts decided what bias to
build in. This chapter turns the lens around and reads the bias back out of
a model that has already learned. It is the empirical payoff of the book's
third axis, implementation realization: once gradient descent has run, which
of the architecturally permitted solutions did it actually find, and can we
recover that solution from the weights?

The $M = 8$ model is small enough to answer cleanly. It has 17{,}282
parameters and the task has $2^{11} = 2048$ possible inputs, so we can
describe what every component does and do it from the trained weights and
attention patterns directly. That is mechanistic interpretability at its
cleanest: a task with known structure, a model that solves it, and a
hypothesis specific enough to falsify. The first eight sections carry out
that dissection and find the predicted circuit, confirmed by ablations. They
also place the circuit in the literature, as a near-cousin of the induction
head and a minimal instance of the mechanism behind in-context learning.

Then the chapter does the part it did not set out to do. The same task at
$M = 32$ needs a third layer, and when we point the same tools at a model
that solves it, the clean circuit is gone. The attention maps look
illegible, and two hypotheses written before the analysis are refuted by the
data. A causal probe then shows that the model is nonetheless a perfect
pointer, and a causal trace recovers a mechanism as clean as the $M = 8$
one, a mechanism the attention maps simply could not surface. The lesson in
one sentence: attention weight is not information flow, and the tool, not
the model, decides what is legible. This is the book's flagship
honest-empiricism result, and it is the largest single section in the book.

A word on what is being dissected, because the chapter studies two distinct
models in two frameworks and the distinction matters. Sections 1 through 8
analyze the from-scratch NumPy model of \Cref{ch:transformer}: two layers,
one head, $d_{\text{model}} = 32$, the file
\texttt{examples/pointer\_interp.py}. It retrains cheaply on each run, so its
exact weights move a little between runs while the circuit it learns does
not. Section 9 analyzes a separately trained, deliberately introspectable
PyTorch model: three layers, four heads, $d_{\text{model}} = 128$, attention
written out by hand so every pattern is readable, the file
\texttt{examples/pytorch/pointer\_interp\_deep.py}. This is \emph{not} the
Chapter~6 scaling-sweep model, which used a library attention layer; it is a
fresh model built to be read, loaded from a committed checkpoint so its
numbers are fixed. The two models solve the same kind of task at different
scales, but their accuracies are not comparable across the framework
boundary, and the chapter never compares them as numbers. What carries
across is the method and the lesson.

\input{chapters/ch07/01-setup-and-circuit}
\input{chapters/ch07/02-m8-analysis}
\input{chapters/ch07/03-induction-and-icl}
\input{chapters/ch07/04-scaling-the-lens}
\input{chapters/ch07/05-closing}
\input{chapters/ch07/bib-notes}
