% DEFINED:   sec:attention-lookup, sec:pointer-task, sec:one-layer-limit
% RESOLVED:  ch:foundations (backward into Part I)
% FORWARD:   (none)

\section{Attention as a lookup}\label{sec:attention-lookup}

Scaled dot-product attention is the one operation the whole chapter turns on.
For a query matrix $Q$, a key matrix $K$, and a value matrix $V$, it computes
\[
  \mathrm{Attention}(Q, K, V)
  = \mathrm{softmax}\!\left(\frac{Q K^{\top}}{\sqrt{d_k}}\right) V .
\]
Three steps run in sequence. Take the pairwise dot products between every
query and every key. Turn each query's row of scores into a distribution over
positions with a softmax. Return, for each query, the average of the values
weighted by that distribution. Read operationally, that is exactly what a
content-addressable memory does.

\begin{itemize}
\item The query at a position encodes what that position is looking for.
\item The key at each past position encodes what is stored there to be
matched against.
\item The value at each past position encodes what would be returned if that
position is the one selected.
\item The softmax picks, softly, the past position whose key best matches the
query, and the output is that position's value.
\end{itemize}

When the query is the encoding of an address, ``return position $k$,'' and the
keys encode positions, the softmax sharpens to a near one-hot spike on the
addressed position, and its value is read out. That is a pointer dereference,
done with differentiable parts so it can be learned by gradient descent.

This is where the book's pure-Python through-line ends. The operators below
are hand-derived NumPy from \texttt{examples/transformer.py}, vectorized
because pure-Python loops over the $T \times T$ score matrix are not
tractable, but the derivation is the same one \Cref{ch:foundations} carried
for the multilayer perceptron. Here is the core of the forward pass, lifted
from the multi-head attention class and shown for a single head: project the
input into queries, keys, and values, score, mask out the future so a
position can only read the past, softmax, and read out the weighted values.

\begin{lstlisting}
# Q, K, V: (T, d_k) projections of the input x.
scale  = 1.0 / math.sqrt(d_k)
scores = (Q @ K.T) * scale                          # (T, T) match scores
mask   = np.tril(np.ones((T, T)))                   # lower triangular
scores = np.where(mask > 0, scores, -1e9)           # forbid attending ahead
attn   = softmax(scores, axis=-1)                   # (T, T) distribution
out    = attn @ V                                   # (T, d_k) read-out values
\end{lstlisting}

The masking step is what makes this a causal read: position $t$ may match
against positions $0$ through $t$ and no further, so the lookup at the final
position can reach every earlier position but not the answer it is trying to
predict. Whether attention \emph{succeeds} at a given lookup depends on
whether the address it needs is actually present in the query, and that is
where depth enters.

\section{A pointer-dereferencing task}\label{sec:pointer-task}

To test the claim we need a task whose structure \emph{requires} a
content-addressable read, with no shortcut a fixed pattern could exploit. The
task is deliberately small so the structure stays visible. Each example is a
12-bit sequence with three fields:
\[
  \underbrace{m_0\, m_1\, \cdots\, m_7}_{\text{8 memory bits}}\;
  \underbrace{a_0\, a_1\, a_2}_{\text{3 address bits}}\;
  \underbrace{y}_{\text{target}} .
\]
The eight memory bits are random. The three address bits are the binary
encoding of an index $a \in \{0, 1, \ldots, 7\}$, most significant bit first.
The target is $y = m_a$, the memory bit at the addressed position. There are
$2^{11} = 2048$ possible inputs. We draw 20{,}000 training examples with
replacement, so duplicates abound, and 2000 held-out examples for evaluation.
The model reads the first eleven bits and predicts the twelfth. The generator
is \texttt{make\_variant1} in \texttt{examples/simple\_pointer\_dgp.py}:

\begin{lstlisting}
def make_variant1(num_examples, M=8, A=3, seed=0):
    """[M memory bits] [A address bits] [1 target = memory[addr]]."""
    rng = np.random.default_rng(seed)
    memory = rng.integers(0, 2, size=(num_examples, M))
    address_int = rng.integers(0, M, size=num_examples)
    # Encode the address as A binary bits, most significant first.
    address_bits = np.zeros((num_examples, A), dtype=np.int64)
    for j in range(A):
        address_bits[:, j] = (address_int >> (A - 1 - j)) & 1
    target = memory[np.arange(num_examples), address_int][:, None]
    sequences = np.concatenate([memory, address_bits, target], axis=1)
    return sequences, target.squeeze(-1)
\end{lstlisting}

This is the smallest task that requires content-addressable memory. To get
the answer the model has to learn three steps: recognize that the bits at
positions 8, 9, 10 are an address, decode those three bits to an integer $a$,
and use $a$ to read memory position $m_a$. The third step is the dereference,
and it is dynamic: \emph{which} memory position must be read depends on the
\emph{content} of the address bits, which are only known at inference time. No
fixed wiring can stand in for it, which is exactly what makes the task a clean
probe of the prior.

\section{Why one attention layer is not enough}\label{sec:one-layer-limit}

A subtle but load-bearing point: a one-layer, one-head transformer
\emph{cannot} solve this task, and the reason is structural rather than a
matter of capacity.

The prediction is made at the last input position, position 10, using the
output of attention there. The query at position 10, before attention, is
computed from the input at position 10 alone: the single address bit $a_2$,
plus a positional encoding. It does not see $a_0$ or $a_1$, which sit at
positions 8 and 9. The attention layer then forms a weighted sum of past
values with weights set by $\mathrm{softmax}(Q_{10} \cdot K_i)$. Since
$Q_{10}$ is fixed once the input at position 10 is fixed, those weights are
not a function of the full address. The layer can learn a fixed attention
pattern, always read memory position 3, say, but it cannot route to a
position whose identity depends on bits it never sees.

Two layers can. The first layer lets the hidden state at position 10 pull in
the address bits from positions 8 and 9 through attention. The second layer's
query at position 10 is then computed from a hidden state that contains all
three address bits, so it can route dynamically to the addressed memory cell.
This is the precise sense in which transformer depth equals the number of
composed lookups: each additional layer adds one more stage of dynamic
routing on top of the information the previous layers have already gathered.
The next section puts the argument to an experiment.
