% DEFINED:   sec:pe-scale-fix, sec:partial-confirmation, tab:depth-transition
% RESOLVED:  (none new)
% FORWARD:   (none)

\section{The fix: positional-encoding scale}\label{sec:pe-scale-fix}

The audit pinned down why the choice of positional encoding mattered so much.
The \texttt{Embedding} layer initializes its weight matrix at scale $0.02$,
the standard default tuned for large vocabularies. With a two-token bit
vocabulary the input embeddings stay near that $0.02$ scale, while the
sinusoidal positional encoding has entries of magnitude about $1$. So at the
input, where embedding and position are added together, the content signal is
roughly $35$ times smaller than the positional signal. The model has to learn
to discount the positional swamping before it can route on content. At $M=8$
that finishes in time. At $M \geq 16$ the optimization plateaus before the
embeddings have been rescaled to anything competitive with the encoding.

There are two clean fixes, and both target the same content-versus-position
\emph{scale balance} at the network's input.

\begin{enumerate}
\item \textbf{Rescale the embedding} so content and position start at
comparable magnitudes. The original transformer paper does exactly this: it
multiplies the embedding output by $\sqrt{d_{\text{model}}}$ before adding the
positional encoding. A crude $25\times$ rescaling of the embedding weights
alone lifts $M=16$ from 0.67 to 0.79 with no other change, which was checked
during the audit.
\item \textbf{Use a learned positional encoding}, initialized at the same
small $0.02$ scale as the embedding. Now content and position both start small
and the model can grow them together as training proceeds. This is what takes
the lookup task to 100 percent in the architecture sweep of
\Cref{sec:four-hypotheses}.
\end{enumerate}

Modern transformer libraries use either learned or rotary position
representations, both of which sidestep the fixed-magnitude
sinusoidal-against-tiny-embedding problem. At production scale, with large
vocabularies and well-tuned initialization and warmup, the issue mostly
disappears; at this small experimental scale it dominates. The multi-head
result fits the same story. With several heads, at least one can find a
query-key geometry that compensates for the input-scale imbalance, redundancy
a single head does not have, which is why four heads recover 98.3 percent
where one head reaches only 0.747.

\section{Verification: the bias is partially confirmed}\label{sec:partial-confirmation}

With the fix in hand we rerun the scaling experiment, changing only the
positional encoding (sinusoidal versus learned, both at the same $0.02$ init
scale) and holding everything else at the baseline: two layers, one head,
$d_{\text{model}}=64$, 6000 iterations.

\begin{center}
\begin{tabular}{@{}rrr@{}}
\toprule
$M$ & Sinusoidal PE & Learned PE \\
\midrule
16 & 0.747 & \textbf{1.000} \\
24 & 0.687 & 0.605 \\
32 & 0.664 & 0.645 \\
\bottomrule
\end{tabular}
\end{center}

At $M=16$ the fix is decisive: learned positional encoding solves the lookup
task to 100 percent, exactly as the content-addressable-memory theory predicts
once the input-scale obstacle is removed. At $M \geq 24$, though, the same fix
does not carry. Both encodings stay near chance, so the scale mismatch that
dominated at $M=16$ is one barrier among several, and another kicks in at
larger $M$. That nuance, frustrating to hit but honest to report, sharpens
this chapter's lesson rather than softening it. The architectural inductive
bias is \textbf{necessary but not sufficient}. Whether a network realizes the
bias depends on a co-evolved set of choices the theory glosses over:
initialization, positional encoding, supervision shape, optimizer state,
learning-rate schedule, width, depth, data volume, and training time.

To see how far the standard production knobs close the gap, we ran a
scaled-down production recipe (the file is
\texttt{examples/pointer\_kitchen\_sink.py}): $d_{\text{model}}=128$ instead
of 64, four heads instead of one, learned positional encoding at small init,
learning-rate warmup over 500 iterations, and 8000 iterations, with the same
two layers and the same 20{,}000 examples.

\begin{center}
\begin{tabular}{@{}rrrr@{}}
\toprule
$M$ & Sinusoidal (baseline) & Learned PE alone & Kitchen-sink recipe \\
\midrule
16 & 0.747 & \textbf{1.000} & \textbf{1.000} \\
24 & 0.687 & 0.605 & \textbf{0.9975} \\
32 & 0.664 & 0.645 & 0.680 \\
\bottomrule
\end{tabular}
\end{center}

The $M=24$ row is the headline. The combined recipe lifts accuracy from
0.605 to 0.9975, a near-complete solve at a scale that learned encoding alone
could not touch. The bias was present in the architecture the whole time; the
production defaults are what let it become a learned routine. At $M=32$ the
recipe barely helps, reaching only 0.680, so we chased that case down. Three
knobs were left: more iterations, more width, more depth. We tried each in
isolation. To run a proper sweep without waiting hours per configuration, the
same architecture was reimplemented in PyTorch.

A caution that matters, and that the table below carries: the PyTorch model
uses different initialization and different internals from the from-scratch
NumPy code, so its absolute accuracies are \emph{not} comparable to the NumPy
numbers above. The PyTorch results are read only against each other. With
that wall in place, the result is clean. More iterations alone do not solve
$M=32$: the NumPy kitchen-sink run to 30{,}000 iterations reached only 0.7625,
and a PyTorch two-layer model run to 40{,}000 iterations sat at chance-plus
with a loss curve flat from iteration 2000 on. More width alone does not
solve it either: doubling to $d_{\text{model}}=256$ with eight heads, four
times the parameters at two layers, stayed indistinguishable from chance-plus.
Depth does. Holding width, head count, data, and seed fixed and varying only
the layer count over a 40{,}000-iteration budget:

\begin{table}[t]
  \centering
  \begin{tabular}{@{}rrrrl@{}}
  \toprule
  Layers & Parameters & Loss at 40k & Test accuracy & Loss curve \\
  \midrule
  2 & 270k & 0.6463 & 0.6160 & flat, never transitions \\
  3 & 403k & 0.1154 & \textbf{0.9460} & transitions at iter ${\sim}7000$ \\
  \bottomrule
  \end{tabular}
  \caption{The $M=32$ depth transition, run in PyTorch with only the layer
  count varying (the controlled comparison from
  \texttt{examples/RESULTS.md}). The two-layer model is flat for the whole
  budget; the three-layer model phase-transitions near iteration 7000 and is
  still descending when stopped. The PyTorch model's initialization and
  internals differ from the NumPy model used elsewhere in this chapter, so
  these numbers are read only against each other, not against the NumPy
  tables.}
  \label{tab:depth-transition}
\end{table}

\Cref{tab:depth-transition} is decisive within its frame: at $M=32$, depth is
the binding knob, and neither width nor training time substitutes for it. It
is tempting to package this as a tidy law, that depth equals the number of
address bits the model must compose, so larger $M$ needs more layers. Resist
it, because the data refute the clean version. $M=24$ and $M=32$ have the same
address width, since $\lceil \log_2 24 \rceil = \lceil \log_2 32 \rceil = 5$,
yet two layers solved $M=24$ to 0.9975 and cannot solve $M=32$ at all. The
address-decode work is identical across the two. What differs is the
\textbf{lookup fan-out}, the number of memory cells the second-stage query
must discriminate among, 24 versus 32, together with the fact that $M=32$
fills the entire five-bit address space while $M=24$ uses only 24 of the 32
codes. The bottleneck at $M=32$ is the precision of the one-of-$M$ selection,
not the depth of the address decode. The honest claim is the controlled-
experiment one, that at $M=32$ depth solves where width and time do not, and
not a closed-form depth law.

This is the scaling-law picture in miniature. Capacity, depth, data, and
iterations co-scale, and the binding constraint moves as the task hardens. At
$M=8$ the binding constraint was nothing; any two-layer model solves it. At
$M=16$ it was positional-encoding scale. At $M=24$ it was the combined
production recipe. At $M=32$ it is depth. Each regime hides a different
bottleneck, and finding it is the whole game.
