% DEFINED:   ch:transformer
% CONTAINS:  section/figure/table labels live in the \input'd files below
%            (sec:attention-lookup, sec:pointer-task, sec:one-layer-limit,
%            sec:exp-single-lookup, fig:ch06-grokking, sec:multihead,
%            sec:exp-multihop, sec:scaling-puzzle, fig:ch06-scaling,
%            sec:four-hypotheses, sec:pe-scale-fix, sec:partial-confirmation,
%            tab:depth-transition, sec:three-axes, sec:richer-dgp)
% RESOLVED:  ch:foundations, sec:inductive-biases, sec:parallel-axes, ch:cnn,
%            ch:rnn, ch:fixed-context-lm (all backward into Part I and earlier
%            Part II)
% CROSS-CHAPTER FORWARD (prose-only, no \Cref): Chapter 7 (interpretability:
%            reverse-engineering this trained transformer, induction heads,
%            in-context learning) and Chapter 8 (reinforcement learning; the
%            heads-as-bias frame). Neither is drafted yet; stays prose.

\chapter[Attention as Content-Addressable Memory]{Attention as
Content-Addressable Memory}%
\label{ch:transformer}

\Cref{ch:cnn,ch:fixed-context-lm,ch:rnn} each committed to a structural
prior about \emph{which positions matter}. The convolutional network shared
one kernel across space and bet on locality. The fixed-context model fixed a
window and bet that the recent past was enough. The recurrent network shared
one cell across time and squeezed the whole past through a single state. The
transformer makes a stranger and more flexible commitment. It does not
decide in advance which positions matter. It commits to a \emph{mechanism}
for deciding, from the data, position by position, on the fly.

The architectural claim of this chapter fits in one sentence. An attention
head is a learned content-addressable lookup: a query matches against keys
to retrieve values, exactly as a pointer dereference reads memory by
address. Stacking attention layers composes lookups into multi-hop reads.
That is the entire structural prior of a transformer, and the rest of the
chapter is the unpacking of it.

To make the claim falsifiable we build a synthetic data-generating process
whose structure \emph{requires} content-addressable memory, train small
transformers on it, and watch attention learn to do what its mechanism is
built for. The task has a single-lookup variant and a multi-hop variant, so
it can test both halves of the claim: that one attention layer is a lookup,
and that depth is the number of composed lookups.

The chapter does two things. The first half states the task, derives why one
attention layer cannot compose two lookups, and confirms the depth argument
at small scale. The second half is the part the chapter did not set out to
do. When we scale the task by enlarging the memory, the transformer
suddenly loses to a brute-force multilayer perceptron, and we have to debug
why. The debugging is where the chapter earns its real lesson, and where the
book's third axis appears: the right architecture is necessary but not
sufficient, and whether gradient descent realizes the bias-permitted
solution depends on positional-encoding scale, initialization, head count,
and depth.

This is also where the book's pure-Python through-line ends. The
convolutional, fixed-context, and recurrent models were
\texttt{scratchnn}, model and experiment, standard library only. The
transformer here is hand-derived NumPy. Every backward pass is still
written out by hand, the way \Cref{ch:foundations} wrote out backpropagation
for the multilayer perceptron, but the operators are vectorized in NumPy
because pure-Python loops over the $T \times T$ attention matrix are not
tractable. The listings in this chapter come from that NumPy implementation,
\texttt{examples/transformer.py}. The chapter says plainly where the switch
happens and walls off the one place a second framework appears: a scaling
sweep run in PyTorch, cited as a table, with the warning that its internals
differ from the NumPy model so its numbers are read only against each other.

\input{chapters/ch06/01-lookup-and-task}
\input{chapters/ch06/02-depth-and-multihead}
\input{chapters/ch06/03-scaling-and-investigation}
\input{chapters/ch06/04-fix-and-verification}
\input{chapters/ch06/05-three-axes-and-appendix}
\input{chapters/ch06/bib-notes}
