% DEFINED:   sec:trichotomy, sec:rl-reduction, sec:mdp
% RESOLVED:  (none; this file makes no \Cref)
% FORWARD:   sec:credit-assignment, sec:reinforce (later in this chapter)

\section{The trichotomy}\label{sec:trichotomy}

Three standard learning paradigms, with the boundary that matters drawn at
the right place.

\begin{itemize}
\item \textbf{Supervised.} Given $(x, y)$ pairs, learn $f_\theta$ such that
$f_\theta(x) \approx y$. One gradient per example. The loss tells you the
answer was wrong, \emph{and how to be less wrong on this exact input}.
\item \textbf{Unsupervised.} Given $x$ alone, learn structure: density,
clusters, manifolds, factors. There is no target.
\item \textbf{Reinforcement learning.} Given an environment and a scalar
reward, learn a policy $\pi_\theta(a \mid s)$ that maximizes expected return.
The signal is one number per trajectory, not a label per example.
\end{itemize}

Self-supervised learning, the engine of every modern language model, is not
a fourth paradigm. Next-token prediction, masked-patch reconstruction, and
contrastive pretraining all manufacture their own labels from unlabeled
data, and once the label exists the objective is ordinary supervised
learning. It is mathematically identical and practically distinct only
because the data supply is effectively unbounded, which is what changes the
scaling story. A transformer trained on ten terabytes of web text is running
the same loss minimization as a one-layer softmax classifier on a table of
digits, with more data and a heavier model. Self-supervised learning is
supervised learning that solved its label problem.

So the boundary that matters is not ``do we have labels'' but ``do we have a
\emph{per-example} gradient signal.'' Supervised and self-supervised learning
both do. Reinforcement learning does not, and that single absence is what the
rest of the chapter is about.

\section{The reduction loses what makes RL hard}\label{sec:rl-reduction}

A common move flattens the distinction: supervised learning, the argument
goes, is just reinforcement learning with reward $r = -\mathrm{NLL}(y \mid
x)$ at every step. This is technically true. The reduction packages each
$(x, y)$ pair as a one-step episode in which the agent takes action $\hat y$,
the environment immediately reveals $y$, and the reward is the negative
log-likelihood of $y$ under $\pi_\theta(\cdot \mid x)$. Maximizing expected
reward is then maximum likelihood, and the whole supervised apparatus
reappears as a degenerate special case.

What the reduction quietly discards is everything that makes reinforcement
learning hard:

\begin{itemize}
\item \textbf{Delayed reward.} Here the reward arrives at the same step as
the action. In general it can arrive many steps later, or only at the end.
\item \textbf{Sparse reward.} Here every action has a reward. In general most
actions return nothing, and a single number arrives at the end of a long
trajectory.
\item \textbf{Exploration.} Here the action $\hat y$ and the gradient of the
reward with respect to it are directly available; there is no need to try the
alternatives. In general the agent only ever sees the reward of the action it
actually took.
\item \textbf{Off-policy correction.} Here the data and the policy are the
same thing. In general there is a real distinction between the behavior that
generated the data and the target being improved.
\end{itemize}

Each of those is a hard part of reinforcement learning erased. The genuine
difficulty underneath all of them is \textbf{temporal credit assignment}. A
trajectory $\tau = (s_0, a_0, r_1, s_1, a_1, r_2, \ldots, s_T)$ produces a
single return $R(\tau) = \sum_t r_t$. The learner took, say, two hundred
actions. Which of them were responsible for the return being high or low?

REINFORCE, advantage estimation, temporal-difference learning, generalized
advantage estimation, eligibility traces, hindsight relabeling, off-policy
corrections: every named technique in the field is a piece of that
credit-assignment machinery. Supervised learning never needs any of it,
because the label tells you precisely which input produced which loss, and
the chain rule walks the blame backward through the network. The
reinforcement-learning loss surface, by contrast, is computed \emph{after}
the actions are committed, and the trajectory probability is a product over
many policy decisions of which only some mattered. The reframing earns its
keep because it lets a reader who has only done supervised work see
\emph{why} reinforcement learning is hard, not merely \emph{that} it is.

\section{The MDP framing}\label{sec:mdp}

The standard formalism is the \textbf{Markov decision process}. An agent and
an environment interact in a loop. At step $t$ the agent observes a state
$s_t$, takes an action $a_t \sim \pi_\theta(\cdot \mid s_t)$, and the
environment returns a reward $r_{t+1}$ and a next state $s_{t+1} \sim p(\cdot
\mid s_t, a_t)$.

The \textbf{Markov property} is the assumption that the transition
distribution $p(s', r \mid s, a)$ depends only on the current $(s, a)$, not on
the prior history. The ``state'' is, by definition, whatever you need to
remember for that to hold.

The \textbf{discount} is a choice of $\gamma \in [0, 1)$. The discounted
return from step $t$ is
\[
  G_t = \sum_{k=0}^{\infty} \gamma^k\, r_{t+k+1}.
\]
The discount is geometric weighting on the future. At $\gamma = 0$ the agent
is myopic and only the next reward matters; as $\gamma \to 1$ a reward now and
a reward in a thousand steps are valued equally. In practice $\gamma \in [0.9,
0.999]$ is typical, and the effective horizon is about $1 / (1 - \gamma)$
steps.

The \textbf{objective} is to maximize the expected return
\[
  J(\theta) = \mathbb{E}_{\tau \sim \pi_\theta}\!\left[\, R(\tau) \,\right],
\]
where $\tau = (s_0, a_0, r_1, s_1, \ldots)$ is a trajectory sampled by
following $\pi_\theta$ in the environment and $R(\tau) = \sum_{t \ge 0}
\gamma^t r_{t+1}$. Two derived quantities organize most algorithms, the
state-value and action-value functions:
\[
  V^\pi(s) = \mathbb{E}_\pi[\,G_t \mid s_t = s\,],
  \qquad
  Q^\pi(s, a) = \mathbb{E}_\pi[\,G_t \mid s_t = s,\, a_t = a\,].
\]
A policy is a conditional distribution over actions given states; a return is
the discounted sum of rewards along the trajectory the policy induces; we
want the policy whose induced trajectories have the highest expected return;
and $V^\pi$ and $Q^\pi$ are the natural quantities to estimate along the way.
The next section asks how to turn samples of $R(\tau)$ into a gradient on
$\theta$ at all.
