% DEFINED:   sec:credit-assignment, sec:reinforce, sec:reinforce-bridge
% RESOLVED:  sec:softmax (Ch1, the softmax cross-entropy gradient),
%            ch:output-heads (heads-as-priors), ch:foundations (the MLP)
% FORWARD:   sec:gridworld (later in this chapter)

\section{Credit assignment, three families}\label{sec:credit-assignment}

Supervised learning gives every example its own loss, and the chain rule pins
the blame precisely. Reinforcement learning gives one return per episode, or a
few sparse rewards along the way, with a sequence of many actions between
them, and the question of which action mattered is genuinely open. There are
three families of answers, each a different bet about how to split the credit.

\begin{itemize}
\item \textbf{Value-based (and Monte Carlo at the extreme).} Estimate how good
states or actions are. The Monte Carlo version waits until the episode ends
and attributes the entire realized return $G_t$ to every action taken,
optionally discounted by recency: high variance, since one trajectory is one
sample, but unbiased, since the return is what actually happened.
Temporal-difference methods do not wait; they use a learned value estimate
$V(s_{t+1})$ as a bootstrap target for $V(s_t)$ through the Bellman equation,
which lowers variance at the cost of a bias from the accuracy of $V$.
\item \textbf{Policy-gradient.} Skip the per-trajectory credit-assignment
question entirely. Estimate the gradient of $J(\theta)$ as an expectation over
trajectories, average many samples, and let the averaging do the work of
variance reduction. This is the family the chapter develops, because the
bridge to supervised cross-entropy is exact and the derivation is short.
\item \textbf{Model-based.} Learn the environment's dynamics $p(s' \mid s, a)$
and plan against the learned model, paying the cost of fitting it in exchange
for being able to imagine rollouts rather than execute them.
\end{itemize}

Actor-critic, $n$-step returns, generalized advantage estimation, and
eligibility traces are interpolations between these: bias-variance trade-offs
on the shape of the credit signal.

\section{REINFORCE, derived}\label{sec:reinforce}

Parameterize the policy as $\pi_\theta(a \mid s)$. The mental image is a
neural network outputting a softmax over discrete actions, or the parameters
of a continuous distribution; the gridworld below uses the discrete-softmax
form, with the same multilayer perceptron of \Cref{ch:foundations} behind the
head. The objective is the expected return under the policy-induced trajectory
distribution,
\[
  J(\theta) = \mathbb{E}_{\tau \sim \pi_\theta}[R(\tau)]
            = \sum_\tau p_\theta(\tau)\, R(\tau),
\]
with $p_\theta(\tau)$ the probability of sampling trajectory $\tau$. We want
$\nabla_\theta J(\theta)$.

The load-bearing identity is the \textbf{log-derivative trick}. For any
parameterized distribution $p_\theta$,
\[
  \nabla_\theta p_\theta(\tau) = p_\theta(\tau)\,
      \nabla_\theta \log p_\theta(\tau),
\]
which is nothing more than $\nabla \log x = \nabla x / x$ rearranged. It turns
the gradient of a probability, which is awkward because probabilities are
normalized, into the gradient of a log-probability, which factors cleanly,
times the probability itself, which becomes an expectation as soon as we
sample. Applying it to the objective,
\[
  \nabla_\theta J(\theta)
    = \sum_\tau \nabla_\theta p_\theta(\tau)\, R(\tau)
    = \mathbb{E}_{\tau \sim \pi_\theta}\!\left[
          \nabla_\theta \log p_\theta(\tau)\, R(\tau)
      \right].
\]

Now factor the trajectory probability. A trajectory is sampled by alternating
policy choices and environment transitions:
\[
  p_\theta(\tau) = p(s_0) \prod_{t \ge 0}
      \pi_\theta(a_t \mid s_t)\, p(s_{t+1} \mid s_t, a_t).
\]
Take the log, turning the product into a sum, and differentiate with respect
to $\theta$:
\[
  \nabla_\theta \log p_\theta(\tau)
    = \sum_t \nabla_\theta \log \pi_\theta(a_t \mid s_t).
\]
The initial-state distribution $p(s_0)$ and the transition kernel $p(s_{t+1}
\mid s_t, a_t)$ do not depend on $\theta$, so their gradients vanish. Only the
policy factors survive. This is the structural fact that makes model-free
reinforcement learning possible: you can estimate the policy gradient without
ever knowing the environment's dynamics.

Substituting back, and using the causality argument that a reward earned
before step $t$ cannot have been influenced by the action $a_t$ (so its
contribution to that step's gradient averages to zero), the
\textbf{reward-to-go} form of REINFORCE drops out:
\[
  \boxed{\;\nabla_\theta J(\theta)
    = \mathbb{E}_{\tau \sim \pi_\theta}\!\left[
        \sum_t \nabla_\theta \log \pi_\theta(a_t \mid s_t)\, G_t
      \right].\;}
\]
In words: averaged across sampled trajectories, the gradient of the expected
return is the sum over time steps of the log-likelihood gradient of the action
taken, each weighted by the return earned from that step onward. Sample one
trajectory, compute that sum, and you have an unbiased Monte Carlo estimate of
$\nabla_\theta J$; step along it with SGD.

\textbf{Variance reduction by a baseline.} Subtracting any state-dependent
baseline $b(s_t)$ from $G_t$ leaves the expectation unchanged but typically
reduces variance:
\[
  \nabla_\theta J(\theta)
    = \mathbb{E}\!\left[
        \sum_t \nabla_\theta \log \pi_\theta(a_t \mid s_t)\,
        \bigl(G_t - b(s_t)\bigr)
      \right].
\]
The expectation is unchanged because $\mathbb{E}_a[\nabla_\theta \log
\pi_\theta(a \mid s)] = 0$ for every $s$, so multiplying by a state-dependent
constant and summing contributes nothing in expectation. How much the variance
drops depends on how strongly $b(s_t)$ correlates with $G_t$; a learned value
estimate $V_\phi(s_t)$ is the standard choice, and $G_t - V_\phi(s_t)$ is then
an estimate of the advantage $A^\pi(s, a)$. That is the half-step from
REINFORCE to actor-critic. The gridworld example uses a simpler scalar
baseline, a moving average of recent episode returns, which is enough to make
the gradient useful at that scale.

\section{The bridge: REINFORCE is weighted cross-entropy}\label{sec:reinforce-bridge}

This is the chapter's payoff, and it is a derivation, not an analogy. Look at
the per-step term in the REINFORCE update, written as a loss to descend (the
minus sign is so that descending on it ascends on expected return):
\[
  \ell_t(\theta) = -\,\log \pi_\theta(a_t \mid s_t)\, \cdot\, G_t.
\]
The whole claim is that the first factor is, identically, the supervised
softmax cross-entropy gradient of \Cref{sec:softmax}. Make the policy concrete
the way the gridworld does. The network produces a vector of logits $z =
g_\theta(s_t) \in \mathbb{R}^K$ over the $K$ actions, and the policy is their
softmax,
\[
  \pi_\theta(a \mid s_t) = \mathrm{softmax}(z)_a
    = \frac{e^{z_a}}{\sum_{j} e^{z_j}}.
\]
Treat the action actually taken, $a_t$, as a one-hot label, and the
log-policy of that action is exactly the negative cross-entropy between that
one-hot and the predicted distribution:
\[
  -\log \pi_\theta(a_t \mid s_t)
    = -\log \mathrm{softmax}(z)_{a_t}
    = \underbrace{\log\!\Bigl(\textstyle\sum_j e^{z_j}\Bigr) - z_{a_t}}
              _{\text{softmax cross-entropy on the label } a_t}.
\]
Differentiate with respect to the logits, the same one-line computation
carried out in \Cref{sec:softmax}. The log-sum-exp term contributes
$\mathrm{softmax}(z)_k = p_k$ in coordinate $k$, and the $-z_{a_t}$ term
contributes $-1$ in coordinate $a_t$ only:
\[
  \frac{\partial}{\partial z_k}\bigl(-\log \pi_\theta(a_t \mid s_t)\bigr)
    = p_k - \mathbf{1}[k = a_t],
  \qquad\text{that is}\qquad
  \frac{\partial \ell_t}{\partial z}\bigg|_{G_t = 1}
    = p - \mathrm{onehot}(a_t).
\]
This is the predicted distribution minus the one-hot of the chosen action: the
\emph{same} expression the supervised softmax head produces, the one
\texttt{SoftmaxCrossEntropy} returns from its backward pass. The full
per-step REINFORCE gradient is just that vector scaled by the return:
\[
  \frac{\partial \ell_t}{\partial z}
    = \bigl(p - \mathrm{onehot}(a_t)\bigr)\, G_t.
\]
Nothing else changes. The chain rule from the logits back through the head and
the multilayer perceptron is identical to the supervised case, because it is
the same network. The only edit to the entire backward pass is a single scalar
multiply per time step.

So policy gradient is supervised softmax cross-entropy on the actions the
agent took, with each per-sample loss reweighted by the return $G_t$ (or by an
advantage). Reading the same identity at three weights tells the whole family
apart:

\begin{itemize}
\item \textbf{Behavioral cloning} is the constant-weight case. Given a dataset
of expert $(s, a)$ pairs, fit $\pi_\theta$ by ordinary softmax cross-entropy.
It is pure supervised learning of a policy.
\item \textbf{REINFORCE} is the same softmax cross-entropy on the actions the
\emph{agent} took, weighted by how good the return turned out to be.
\item \textbf{Actor-critic} is the same softmax cross-entropy weighted by an
advantage, with a learned baseline subtracted from $G_t$.
\end{itemize}

In supervised learning the dataset tells you what the right action was. In
reinforcement learning the return tells you, after the fact, how good the
action you took was, and you weight the gradient accordingly. The loss surgery
is the same; what differs is who supplies the label, you or the environment,
and how the per-sample importance is scored, uniformly or by return. This is
the through-line of the whole book stated at its sharpest. The supervised
toolkit assembled in \Cref{ch:foundations} and \Cref{ch:output-heads}, a
\texttt{Linear} stack, a $\tanh$ nonlinearity, and a softmax cross-entropy
head, is not merely \emph{analogous} to the inner loop of reinforcement
learning. It \emph{is} the inner loop. The gridworld implements this literally:
its update is a softmax over the episode's states, a subtract-the-one-hot
step, and a per-step scale by the return.

\begin{lstlisting}
# The REINFORCE gradient for one episode (examples/reinforce_gridworld.py).
# states: (T, 25) one-hot states; actions: (T,); weights: (T,) = G_t - baseline.
Z2 = H @ self.W2 + self.b2                  # action logits, shape (T, 4)
Z2 = Z2 - Z2.max(axis=1, keepdims=True)
P  = np.exp(Z2); P /= P.sum(axis=1, keepdims=True)   # softmax over actions

dlogits = P.copy()
dlogits[np.arange(T), actions] -= 1.0       # p - onehot(a): the Ch.1 gradient
dlogits *= weights[:, None]                 # scale each step by G_t - baseline
# ... and the rest is the ordinary MLP backward pass through W2, tanh, W1.
\end{lstlisting}

The line \texttt{dlogits[np.arange(T), actions] -= 1.0} is the entire
conceptual content of the chapter. It is \Cref{sec:softmax}'s $p -
\mathrm{onehot}(a)$, evaluated at every step of the trajectory; the line below
it, the scale by the return, is the only thing reinforcement learning adds to
the supervised gradient.
