% DEFINED:   sec:aixi, sec:rl-axis
% RESOLVED:  ch:fixed-context-lm (the Solomonoff north star), ch:cnn, ch:rnn,
%            ch:transformer (architecture axis as policy-architecture choices),
%            ch:output-heads (heads-as-bias: policy = softmax, value = MSE)
% FORWARD:   (none)

\section{AIXI: the theoretical north star}\label{sec:aixi}

Solomonoff induction, the ideal that framed \Cref{ch:fixed-context-lm}, is the
optimal incomputable \emph{predictor}: a Bayesian mixture over all computable
environments, each weighted by its program length under a universal prior. It
is what an idealized agent would do if it could enumerate every Turing
machine, score each by how well it explains the observations seen so far, and
weight its predictions by the prior times the likelihood. Practical language
models, the fixed-context network of \Cref{ch:fixed-context-lm} among them, are
computable approximations to that ideal.

\textbf{AIXI} (Hutter, 2000) is its action-taking sibling. Where Solomonoff
predicts, AIXI \emph{acts}. At each step $t$ with interaction history
$h_{<t}$, AIXI selects
\[
  a_t^{*} = \arg\max_{a_t} \sum_{o_t r_t}
            \xi(o_t r_t \mid h_{<t}\, a_t)\,
            V^{*}(h_{<t}\, a_t\, o_t\, r_t),
\]
where $\xi$ is the Solomonoff prior over environments (a mixture over all
programs that could have generated the observed history), $o_t$ is the next
observation, $r_t$ the next reward, and $V^{*}$ the optimal value function
under that prior, defined by the same recursion. Decoded: AIXI maintains a
posterior over every environment consistent with what it has seen, and at each
step takes the action whose expected future discounted reward, averaged over
that posterior, is highest. The expectation runs over the infinite-horizon
trajectory under the optimal policy in each candidate environment, which is
itself an AIXI-style recursion.

So AIXI is, exactly, Solomonoff induction plus Bayesian decision theory plus
reward maximization, fastened together. Each piece is a component a full
treatment of reinforcement learning would unpack on its own: the Solomonoff
prior is a universal Bayesian prior, computable to no finite degree but the
right ideal for a learner that wants robustness to model misspecification;
Bayesian decision theory says the action maximizing posterior expected utility
is the action a coherent agent should take; and reward maximization sets that
utility to a discounted sum of rewards, which is itself a strong commitment
(the subject of the next section).

\textbf{The parallel.} The supervised arc had Solomonoff induction as the
unachievable but illuminating prediction ideal, with modern language models as
its practical computable approximations. Reinforcement learning has AIXI as the
unachievable but illuminating \emph{action} ideal, with modern algorithms
(REINFORCE, deep $Q$-networks, proximal policy optimization, MuZero, decision
transformers) as its practical computable approximations. In both arcs the
architecture is the computable inductive bias substituted for an incomputable
optimum. You cannot run Solomonoff; you can run a transformer. You cannot run
AIXI; you can run proximal policy optimization. The architecture \emph{is}
where you compromise, and reading AIXI as the target sharpens what each
practical method is doing. $Q$-learning learns a single environment model
implicitly where AIXI averages over all of them; proximal policy optimization
uses a parameterized policy and a clipped surrogate objective where AIXI takes
an exact arg-max over an infinite expectation; decision transformers collapse
policy and value into one sequence-prediction problem where AIXI keeps them
separate. Each compromise can be located as a specific distance from the AIXI
specification.

\section{Inductive bias, the reinforcement-learning axis}\label{sec:rl-axis}

The supervised chapters argued that architecture (\Cref{ch:cnn},
\Cref{ch:rnn}, \Cref{ch:transformer}) and output head (\Cref{ch:output-heads})
are parallel axes of inductive bias. The same frame extends to reinforcement
learning, with more design surface, each piece of which encodes an assumption
about the task.

\begin{enumerate}
\item \textbf{Reward shaping.} The reward function is a prior about the value
landscape. A sparse goal-reward, the gridworld setup above, is minimal-prior:
the agent has to find a successful trajectory before it gets any informative
gradient at all. A shaping reward, a small positive bonus for moving closer to
the goal, injects information about which intermediate states count as
progress. The same lever can mislead if the hint is wrong, which has its own
literature under reward hacking and Goodhart's law. The reward is a teacher,
and the choice of teacher is a prior.
\item \textbf{Policy architecture.} Whether $\pi_\theta$ is a multilayer
perceptron, a convolutional network over pixels, a recurrent network over
partial observations, or a transformer over state-action histories is the
\emph{same} architectural choice the supervised chapters catalogued. The
convolutional locality of \Cref{ch:cnn}, the recurrence of \Cref{ch:rnn}, and
the content-addressable lookup of \Cref{ch:transformer} all carry over intact,
now in service of action selection rather than prediction. The gridworld uses
a multilayer perceptron because its state is small, discrete, and structureless;
Atari from pixels uses a convolutional network; partially observed tasks with
hidden state use a recurrent network or a transformer.
\item \textbf{Exploration strategy.} Epsilon-greedy, an entropy bonus,
intrinsic curiosity, upper-confidence-bound action selection, Thompson
sampling: each is a prior about where useful information lives. Epsilon-greedy
assumes uniform-random is good enough some of the time; curiosity assumes novel
states are valuable; Thompson sampling assumes a posterior over models and
exploits its variance. Different priors, different exploration behavior, and a
real source of difference between algorithms on sparse-reward tasks.
\item \textbf{Algorithm class.} On-policy against off-policy, model-free
against model-based, value against policy against actor-critic: each commits to
a structural assumption about what is cheap, what is reliable, and what
generalizes. Off-policy methods assume trajectories from older policies still
inform the current one; model-based methods assume a learned dynamics model
will repay the cost of fitting it; value-based methods assume the Bellman
recursion is more reliable than direct policy parameterization. The choice is a
prior about the structure of the task and of the learning dynamics.
\end{enumerate}

All of these are inductive biases in exactly the sense that translation
equivariance is for a convolutional network. The heads-as-bias frame of
\Cref{ch:output-heads} carries over cleanly too: a discrete policy is a
categorical (softmax) head, and a value function is an identity-plus-squared-
error (Gaussian) head, the same two output-distribution choices already in
hand, now predicting actions and returns instead of labels. A real
reinforcement-learning system is a stack of priors at every level, reward,
policy architecture, exploration, algorithm class, and matching each to the
task is a separate engineering problem. The pattern from \Cref{ch:output-heads}
through \Cref{ch:transformer} generalizes to reinforcement learning; it just
has more axes.
