% DEFINED:   ch:rl
% CONTAINS:  section/figure labels live in the \input'd files below
%            (sec:trichotomy, sec:rl-reduction, sec:mdp,
%            sec:credit-assignment, sec:reinforce, sec:reinforce-bridge,
%            sec:gridworld, fig:gridworld-learning-curve, sec:aixi,
%            sec:rl-axis, sec:book-close)
% RESOLVED:  ch:foundations, sec:softmax, ch:output-heads,
%            ch:fixed-context-lm, ch:cnn, ch:rnn, ch:transformer,
%            sec:inductive-biases, sec:three-axes
%            (all backward into Parts I-III)
% FORWARD:   NONE. This is the last chapter of the book.
%            It does not reference Chapter 7 (it ties back to the
%            supervised arc, not interpretability); that omission is
%            intentional, not a gap.

\chapter[Reinforcement Learning]{Reinforcement Learning: A Scalar Reward
over Trajectories}%
\label{ch:rl}

Every chapter so far has varied one of two parallel axes. The architecture
moved from the multilayer perceptron of \Cref{ch:foundations} through the
convolutional, recurrent, fixed-context, and attention families of Part~II.
The output head moved through the Bernoulli, categorical, Gaussian, and
Poisson choices of \Cref{ch:output-heads}. Both axes lived entirely inside
\textbf{supervised learning}. The training signal was always a label per
example, and the gradient question was always the same: how do I move
$f_\theta(x)$ closer to this specific $y$?

Reinforcement learning is the paradigm where that signal disappears at the
example level and reappears as a scalar reward, summed along a whole
trajectory. The distinctive hard part is not the absence of labels. It is
\textbf{temporal credit assignment}: a trajectory of two hundred actions
produces one number, and the learner has to work out which of those actions
mattered. Every named technique in reinforcement learning, from Monte Carlo
returns to temporal-difference bootstrapping to generalized advantage
estimation to eligibility traces to hindsight relabeling, is a piece of
credit-assignment machinery. Supervised learning never needs any of it,
because it never has the problem: the label pins the blame on the input
exactly, and the chain rule walks that blame backward through the network.

This chapter does the minimum to draw the line between supervised learning
and reinforcement learning precisely, and then to dissolve part of it. It
sets up the trichotomy and the Markov decision process, derives the
REINFORCE estimator, and then reads the result back as something already
familiar: softmax cross-entropy weighted by return. The per-step gradient is
the very expression of \Cref{sec:softmax}, the predicted distribution minus
the one-hot of the chosen action, now scaled by how good the trajectory
turned out. The supervised toolkit assembled across Parts~I and~II is, quite
literally, the inner loop of reinforcement learning. A real NumPy gridworld
makes that concrete, AIXI names the incomputable north star (mirroring the
Solomonoff ideal of \Cref{ch:fixed-context-lm}), and the closing section
ends the book.

\input{chapters/ch08/01-trichotomy-and-mdp}
\input{chapters/ch08/02-reinforce}
\input{chapters/ch08/03-gridworld}
\input{chapters/ch08/04-aixi-and-axis}
\input{chapters/ch08/05-closing}
\input{chapters/ch08/bib-notes}
