% DEFINED:   sec:bengio-bias, sec:bengio-training
% RESOLVED:  sec:softmax (backward into Part I)
% FORWARD:   sec:bengio-alice (Task 5)

\section{The inductive biases}\label{sec:bengio-bias}

The model makes three architectural commitments, each one a deliberate
bet about how language works.

\textbf{Markov of order $N$.} The next token depends only on the previous
$N$ tokens. Anything older is forgotten by construction, not by a learned
decay. This is the central prior, and it cuts both ways. Choose $N$ too
small and the model cannot represent agreement that spans a longer
distance; choose $N$ too large and the head's first weight matrix grows
linearly in $N$, costing parameters and inviting overfitting. The window
is a hard architectural knob, set before training and fixed.

\textbf{Shared embeddings.} Every token has one embedding vector,
independent of where in the window it appears. This is weight sharing
across the position axis, applied only to the embedding layer. The same
lookup runs at slot $1$, slot $2$, and so on through slot $N$. It is the
same move the convolutional network of \Cref{ch:cnn} made in space, here
made on the identity of a token: a token means the same thing wherever it
sits, so it should be embedded the same way wherever it sits.

\textbf{Position-sensitive head.} The concatenation step is the choice
that gives the model word order. By placing position $1$'s embedding in
slots $1, \ldots, d$ of $\vx$, position $2$'s in slots $d+1, \ldots, 2d$,
and so on, the MLP head learns \emph{position-specific} features over the
window. It can distinguish ``the cat sat'' from ``sat the cat'' because
each token's embedding occupies a different slice of the input, and the
first weight matrix can weight those slices differently.

Put together, this is a \emph{hybrid} prior, and the hybrid is the point.
The embedding layer is position-equivariant: the same token gets the same
vector regardless of position. The head is position-sensitive: it reads
each slot with its own weights. Bengio gets token-identity sharing for
free, one embedding per token, while still using word order through the
head. The two halves pull in opposite directions on the position axis, and
the architecture holds both at once.

This sits inside a larger design space of choices on the same axis, which
\Cref{sec:window-limits} returns to once the experiment has run: a $1$-D
convolution that would share the head's weights across positions too, an
averaging step that would discard order entirely, an attention mechanism
that would learn which positions matter rather than hardwiring it. Bengio's
choice, concatenate and feed an MLP, is the simplest non-trivial point in
that space. It hardwires positional sensitivity into the architecture with
no convolution and no attention machinery.

\section{Training}\label{sec:bengio-training}

Training needs nothing new. There is no backpropagation through time and no
recurrent unrolling. Each forward pass reads one $N$-token window and
predicts one target token; the backward pass runs from the loss back
through the MLP head, then through \texttt{EmbedConcat}, which splits the
gradient on $\vx$ into $N$ chunks of size $d$ and routes each back through
the embedding in LIFO order. The output head is the categorical one from
\Cref{sec:softmax}: a softmax over the $V$ vocabulary logits, cross-entropy
against the true next token. So the whole training call is one line,
\begin{lstlisting}
net.fit(X, Y, epochs=4, lr=0.02, batch_size=1)
\end{lstlisting}
and the \texttt{Network} loop from \Cref{ch:foundations} handles the rest.
Each training example is independent of every other: there is no
inter-example state to carry forward, no \texttt{reset\_cache()} calls
threaded through the loop, no bookkeeping beyond what \texttt{fit} already
does. The training script is as short as logistic regression's was.

A gradient check confirms the new layer composes correctly. Running the
standard \texttt{gradient\_check} on a small \texttt{Network} built on
\texttt{EmbedConcat} and an MLP head matches central finite differences to
a worst relative error on the order of $10^{-9}$, far under the standard
$10^{-4}$ tolerance, regenerated by the chapter's paired notebook
(\texttt{tests/test\_gradients.py} carries the same check). With the math
verified, the only question left is empirical: how well does an
eight-character window predict English?
