% DEFINED:   sec:window-limits, sec:bengio-map, sec:bengio-handoff
% RESOLVED:  ch:cnn (backward into Part II)
% FORWARD (prose-only): Chapter 5 (the recurrent network, "the next
%            chapter"), Chapter 6 (the Transformer). Per Part II decision A,
%            the full Bengio-vs-recurrent comparison is DEFERRED to Chapter 5;
%            this section gives only a forward preview. No \Cref to Ch5/Ch6.

\section{What the bounded window gives up}\label{sec:window-limits}

The prior that makes this model simple is also its ceiling. Conditioning
on exactly the last $N$ tokens means the model cannot use anything older,
ever. If predicting the token at position $t$ depends on a token at
position $t - 100$, an eight-character window has already forgotten it. For
char-level Alice this rarely bites, because character prediction leans
heavily on local structure: morphology, word boundaries, common bigrams
and trigrams all live well inside eight characters. But the limit is real
and it is structural, not a matter of more training. There are two natural
ways to loosen it, and they point at the next two chapters.

\textbf{A recurrent state.} Instead of a fixed window, carry a hidden
state forward and update it one token at a time. The state is a summary of
the entire past, not a slice of it, so in principle nothing is forgotten by
construction. That is the recurrent network, the subject of the next
chapter. It trades the fixed window for an unbounded-in-principle memory,
and it pays for the trade in ways the fixed-context model does not:
training must run sequentially through time, and the gradient has to travel
back through a long product of Jacobians that tends to shrink it toward
zero. The next chapter builds the recurrent cell on the same corpus and the
same vocabulary used here, precisely so the two priors can be set against
each other directly; the head-to-head comparison of the bounded window and
the recurrent state belongs there, where both models are on the table.

\textbf{A convolution in time.} A $1$-D convolution with kernel size $N$
takes the other fork on the position axis. Where this model gives each
position its own slice of the head's first weight matrix, a convolution
shares one learned context detector across all positions and slides it
along the sequence, exactly the weight sharing the spatial convolution of
\Cref{ch:cnn} used, transposed from space to time. The trade is parameter
efficiency against position specificity. The convolution has fewer
parameters, one kernel applied many times, and it is the natural choice
when the same pattern can appear at any offset, as in motif detection or
edge detection in a $1$-D signal. Bengio's concatenation has more
parameters, one set per position-times-hidden-unit pair, but it can learn
that the immediately preceding token matters more than the one $N-1$ back,
or that agreement runs between two particular slots. For language, where
position carries real information and the nearest token is usually the most
informative, paying for position specificity is a reasonable bet.

The three options form one progression along a single question: how much
of the past does the architecture let a prediction reach, and how is that
reach paid for? The fixed window reaches exactly $N$ tokens, cheaply and in
parallel. The convolution reaches a fixed receptive field while sharing
weights across offsets. The recurrent state reaches arbitrarily far in
principle, at the cost of sequential training and a fragile gradient. Each
chapter in this part takes one more step along that line.

\section{Where this sits in the inductive-bias map}\label{sec:bengio-map}

Two axes organize the architectural priors for sequences: how the model
handles position, and how far back its memory reaches. The fixed-context
model occupies a specific corner of that grid, and naming the neighbors
makes the corner legible.

\begin{center}
\begin{tabular}{@{}lll@{}}
\toprule
Architecture & Position handling & Memory horizon \\
\midrule
MLP & no sharing; flat input & full input dimension \\
$1$-D CNN & weight sharing across positions & fixed receptive field \\
\textbf{Fixed-context LM} & shared embedding, position-sensitive MLP
  & \textbf{fixed window $N$} \\
Recurrent network & weight sharing across time; state carries
  & unbounded in principle \\
Transformer & permutation-equivariant plus positional code
  & full sequence \\
\bottomrule
\end{tabular}
\end{center}

The bottom two rows are forward references: the recurrent network is the
next chapter, and the Transformer is the chapter after it. Read top to
bottom, the table is the same loosening of the memory constraint the
previous section described. The fixed-context model sits between the $1$-D
convolution and the recurrent network: position-sensitive at the head
level, where the convolution is not, and capped at a hard memory cutoff of
$N$ tokens, where the recurrent network is not. It has no direct
interaction between positions beyond what the head learns, which is where
the Transformer, with its explicit pairwise attention, will go further
still.

It is the architecture for problems where the relevant context is small
\emph{and you know how small}. When that assumption holds, it is fast,
simple, and easy to train, with none of the sequential-training or
gradient-stability costs the unbounded-memory architectures carry. When
the assumption fails, when the prediction genuinely needs a token from far
outside the window, no amount of training inside the window can recover it,
and the unbounded-context architectures become necessary.

\section{The Transformer is next}\label{sec:bengio-handoff}

The fixed-context model already shows the appeal of dropping recurrence:
no sequential dependency between examples, no gradient threaded back
through time, just an MLP over a window. Its limit is equally clear: the
window is a hard wall, and inside it the positional structure is hardwired
by concatenation order rather than learned. The next two chapters take the
two ways forward. The recurrent network keeps a running state so the reach
is no longer bounded by a fixed $N$. The Transformer keeps the
fixed-context model's no-recurrence design but replaces the hardwired,
position-by-position head with attention, letting every position read every
other with weights learned from the data, and then adds a positional
encoding back in to recover the order that pure attention throws away. That
is the destination this part is walking toward: from a window, to a state,
to a learned content-addressable lookup over the whole sequence. The
recurrent network comes first.
