% DEFINED:   sec:mle-frame, sec:catalogue, tab:head-catalogue
% RESOLVED:  ch:foundations
% FORWARD:   sec:canonical-link (below in ch02), sec:poisson (below in ch02)

\section{The unifying frame}\label{sec:mle-frame}

Every supervised loss in \Cref{ch:foundations} had the same shape:

\[
  L(\theta) = -\frac{1}{N} \sum_{n=1}^{N} \log p_\theta(y_n \mid x_n).
\]

Minimizing the loss \emph{is} maximum likelihood under an assumed
conditional distribution $p_\theta(y \mid x)$. The network parameterizes
that distribution; gradient descent moves the parameters toward higher
likelihood of the observed data.

The output head is the bridge. The network produces a raw vector
$z = f_\theta(x)$. A \textbf{link function} maps $z$ into the natural
parameter (or the mean, or a probability vector) of the assumed
distribution. The \textbf{loss} is the negative log-likelihood of $y$
under that distribution. \Cref{ch:foundations}'s organizing principle,
\emph{the network produces logits, the loss interprets them}, is exactly
this: the link function and the assumed distribution together \emph{are}
the interpretation.

That is the whole story. The rest of this chapter is the unpacking.

\section{The catalogue}\label{sec:catalogue}

Five canonical pairings, each a member of the exponential family with its
\textbf{canonical link} (the link that makes the gradient come out
cleanest, as \Cref{sec:canonical-link} will prove):

\begin{table}[ht]
\centering
\begin{tabular}{llll}
\toprule
Output type         & Link              & Loss          & Assumed distribution      \\
\midrule
Real-valued         & identity          & MSE           & Gaussian (fixed variance) \\
Binary              & logit (sigmoid)   & BCE           & Bernoulli                 \\
Categorical         & softmax           & cross-entropy & Categorical               \\
Count               & log               & Poisson NLL   & Poisson                   \\
Positive continuous & reciprocal        & Gamma NLL     & Gamma                     \\
\bottomrule
\end{tabular}
\caption{Five canonical head pairings. Each row is a different prior about
  what kind of object $y$ is. The first three are what \Cref{ch:foundations}
  already built; the rest of this chapter extends the table.}
\label{tab:head-catalogue}
\end{table}

Each row is a different commitment about what kind of object $y$ is.
Real-valued and unbounded? Gaussian. A 0 or 1? Bernoulli. One of $K$
mutually exclusive classes? Categorical. A non-negative integer event
count? Poisson.

The link maps the network's unconstrained raw output $z \in \mathbb{R}$
into a valid parameter for the distribution: a real mean for Gaussian, a
probability in $[0, 1]$ for Bernoulli, a probability vector on the simplex
for Categorical, a positive rate for Poisson.

This chapter develops Poisson in full and points to Beta and MDN as natural
extensions. The Gamma row follows the same recipe (positive-support
reciprocal link, Gamma NLL) and is a further natural extension; it is
included in the catalogue for completeness but not developed here.
