% DEFINED:   sec:canonical-link, thm:canonical-link, eq:expfam-grad, sec:three-pairings
% RESOLVED:  sec:logistic, sec:softmax, ch:foundations, eq:residual
% FORWARD:   sec:poisson (below in ch02)

\section{The canonical-link theorem}\label{sec:canonical-link}

\Cref{ch:foundations} observed that both BCE and softmax cross-entropy
have gradient $\partial L / \partial z = \mathbf{p} - \mathbf{y}$ with
respect to the logits: the section on logistic regression in
\Cref{sec:logistic} derived $p - y$ for the binary case, and
\Cref{sec:softmax} derived $\mathbf{p} - \mathbf{y}$ for the
multiclass case. \Cref{ch:foundations} called that pattern no
coincidence but did not prove the general statement. Here it is.

Write an exponential family in natural form:
\[
  p(y \mid \eta) = h(y)\exp\bigl(\eta \cdot T(y) - A(\eta)\bigr),
\]
where $\eta$ is the \textbf{natural parameter}, $T(y)$ is the
\textbf{sufficient statistic}, and
$A(\eta) = \log \int h(y) \exp(\eta \cdot T(y))\,dy$
is the \textbf{log-partition function}.
For Bernoulli, $\eta = \log(p/(1 - p))$ is the logit; for Categorical,
$\eta_i = \log p_i$ up to a shared shift; for Gaussian (fixed variance),
$\eta = \mu/\sigma^2$; for Poisson, $\eta = \log \lambda$.

A foundational identity in exponential families: differentiating the
normalization condition $\int p(y \mid \eta)\,dy = 1$ with respect to
$\eta$ yields
\[
  A'(\eta) = \mathbb{E}\bigl[T(y)\bigr].
\]
The derivative of the log-partition function equals the expected sufficient
statistic. Call this $\hat{p}$, the model's prediction in the relevant
sense: for Bernoulli, $\hat{p}$ is the success probability; for
Categorical, the probability vector; for Gaussian, the mean; for Poisson,
the rate.

Now suppose the network's output $z$ \emph{is} the natural parameter,
$\eta = z$. This is the \textbf{canonical link} choice. The negative
log-likelihood is
\[
  -\log p(y \mid z) = -\log h(y) - z \cdot T(y) + A(z),
\]
and its gradient with respect to $z$ is
\begin{equation}\label{eq:expfam-grad}
  \frac{\partial}{\partial z}\bigl[-\log p(y \mid z)\bigr]
    = -T(y) + A'(z) = \hat{p} - y,
\end{equation}
reading $y$ as $T(y)$ (the sufficient statistic; for Categorical, the
one-hot of $y$).

\begin{theorem}\label{thm:canonical-link}
For any exponential-family distribution with its canonical link, the
gradient of the NLL with respect to the network output is the expected
sufficient statistic minus the observed.
\end{theorem}

\begin{proof}
Differentiate $-\log p(y \mid z)$ directly:
$\partial / \partial z \bigl[{-z \cdot T(y) + A(z)}\bigr] = -T(y) + A'(z)$.
By the exponential-family identity, $A'(z) = \hat{p}$. Reading
$T(y) = y$ gives $\hat{p} - y$.
\end{proof}

\noindent The residual form is not an algebraic accident. The gradient is
geometrically the gap between the model's expected value and what was
observed; the form of the response variable does not enter.

\Cref{ch:foundations}'s pattern is one instance. MSE (identity plus
Gaussian) is another: $\partial L / \partial z = z - y$, which is
$\hat{p} - y$ with $\hat{p} = z$. This is the same residual
\Cref{ch:foundations} wrote as $\hat{y} - y$ in \Cref{eq:residual}: the
prediction $\hat{y}$ there is the Gaussian mean, which is exactly the
expected sufficient statistic $\hat{p}$ here. The Poisson pairing in
\Cref{sec:poisson} will be a third.

\section{Chapter 1's three pairings, named}\label{sec:three-pairings}

Brief recaps; the derivations are in \Cref{ch:foundations}.

\textbf{Identity plus Gaussian plus MSE.} Assumed distribution:
$y \mid x \sim \mathcal{N}(\mu(x), \sigma^2)$ with $\sigma^2$ fixed
and absorbed into the constant. Link: identity (the network output $z$
is $\mu$ directly). NLL (dropping constants): $\tfrac{1}{2}(y - z)^2$,
mean squared error. Gradient: $z - y$. Inductive bias: errors are
symmetric, additive, and Gaussian. If the true distribution is
heavy-tailed or skewed, MSE pays for that with biased estimates.

\textbf{Logit plus Bernoulli plus BCE.} Assumed distribution:
$y \mid x \sim \mathrm{Bernoulli}(p(x))$, $y \in \{0, 1\}$. Link:
logit, $\eta = \log(p/(1-p)) = z$; the inverse is sigmoid,
$p = \sigma(z)$. NLL: $-y \log p - (1-y)\log(1-p)$, computed stably
from $z$. Gradient: $p - y$. Inductive bias: the response is binary; a
probabilistic prediction is well calibrated when training converges.

\textbf{Softmax plus Categorical plus cross-entropy.} Assumed
distribution: $y \mid x \sim \mathrm{Cat}(\mathbf{p}(x))$,
$y \in \{0, \ldots, K-1\}$. Link: softmax,
$\mathbf{p} = \mathrm{softmax}(\mathbf{z})$. NLL:
$-\log p_y = \mathrm{logsumexp}(\mathbf{z}) - z_y$ in the stable form.
Gradient: $p_i - \mathbf{1}[i = y]$, i.e.~$\mathbf{p} - \mathbf{y}$
with $\mathbf{y}$ the one-hot. Inductive bias: the classes are mutually
exclusive and exhaustive; the model cannot represent two classes at once.

Three different objects ($y \in \mathbb{R}$, $y \in \{0,1\}$,
$y \in \{0,\ldots,K-1\}$), three different distributions, three
different links. One pattern: $\hat{p} - y$.
