% DEFINED:   sec:parallel-axes, sec:library-additions, sec:handoff
% RESOLVED:  sec:inductive-biases
% FORWARD (prose-only, no \Cref): Chapter 3 (CNN), Part II, Chapter 8 (RL)

\section{Inductive bias, the parallel axis}\label{sec:parallel-axes}

\Cref{sec:inductive-biases} in Chapter~1 named two axes at which the
prior enters a supervised model and previewed the synthesis. This section
completes it.

\textbf{The two axes, stated precisely:}

\begin{itemize}
\item \textbf{Architecture.} A prior over how the network computes
  features: which inputs interact, at what resolution, with what
  sharing. An MLP assumes every input is independent of every other and
  must learn all relationships from data. A convolutional network assumes
  nearby positions matter together and the same detector should fire
  anywhere. A recurrent network assumes the same computation runs at
  every step of a sequence. A Transformer assumes positions interact
  pairwise through learned, content-addressable lookup.
\item \textbf{Output head.} A prior over the distribution of $y$ given
  the network's raw output: which family of distributions is plausible,
  and what the link function maps the unconstrained output into.
\end{itemize}

The two axes are \emph{independent and composable}. Any architecture can
pair with any output head, provided the head's input dimension matches
the architecture's output. The composability is literal: the
\texttt{Network} class takes a list of layers and a loss; the layers are
the body (any architecture), the loss is the head (any output
distribution). Nothing in the body knows about the head, and nothing in
the head knows about the body.

\textbf{The composability matrix.} Four concrete examples that cross
body type with head type:

\begin{itemize}
\item \textbf{CNN body, Poisson head}: pixel-wise photon-counting
  imaging. The CNN imposes locality and translation equivariance; the
  Poisson head encodes that pixel counts are non-negative integers with
  variance equal to mean.
\item \textbf{Transformer body, Categorical head over a vocabulary}: a
  language model. The Transformer imposes pairwise content-addressable
  attention; the Categorical head encodes that the next token is one of
  $K$ mutually exclusive choices.
\item \textbf{MLP body, heteroscedastic Gaussian head}: regression with
  calibrated, input-dependent uncertainty. The MLP makes no spatial or
  sequential assumption; the Gaussian head encodes that the response is
  real-valued with per-input noise.
\item \textbf{CNN body, Bernoulli head per pixel}: binary segmentation.
  The CNN imposes spatial structure; the Bernoulli head encodes that
  each pixel is independently foreground or background.
\end{itemize}

In each case the architecture is a prior about the \emph{function}
(how features compose) and the head is a prior about the
\emph{distribution} (what $y$ is). They multiply: a matched head buys
sample efficiency \emph{on top of} a matched body, not instead of it.

\textbf{What a matched head gives you:}

\begin{itemize}
\item \textbf{Sample efficiency.} Constraints encoded by the link
  (non-negativity via log, boundedness via sigmoid, simplex membership
  via softmax) are free. The network does not spend parameters on
  relearning them from data.
\item \textbf{Calibrated uncertainty.} A two-output Gaussian head
  reports per-input $\sigma$. A Categorical head reports a probability
  vector. Both are usable by downstream decisions without a
  post-processing step.
\item \textbf{Honesty about the data.} A Poisson head says ``I am
  modeling counts.'' That is a statement to the reader of the code, as
  much as a structural constraint on the optimizer.
\end{itemize}

\textbf{What a mismatched head costs you:}

\begin{itemize}
\item Wasted capacity: the network must discover, from data, constraints
  the link could have given for free (non-negativity, boundedness,
  normalization).
\item Pathological predictions in low-density regions: negative counts
  under MSE, point estimates between modes under a unimodal head,
  miscalibrated confidence under a saturated softmax.
\item Misleading uncertainty: MSE gives no uncertainty at all; a
  poorly calibrated softmax is a known failure mode in deployed
  classifiers.
\end{itemize}

This shape is identical to the architecture story in
\Cref{sec:inductive-biases}. Architectural priors and output-head priors
are parallel commitments, each evaluable on the same questions: does the
prior match the data? does it improve sample efficiency when matched?
does it hurt when mismatched?

The two axes interact but do not conflate. Getting both right is a
multiplication, not an addition: a language model is not just a good
architecture (Transformer) or just a good head (Categorical), it is the
combination. The architecture encodes how context matters; the head
encodes what the prediction is. Neither is optional.

\section{Library additions}\label{sec:library-additions}

\texttt{src/scratchnn/neural\_net.py} contains the following additions
relevant to this chapter. All fit the \texttt{Loss} interface
(\texttt{value}, \texttt{grad}, \texttt{probs}); all are standard-library
only; all have a \texttt{gradient\_check} case.

\begin{itemize}
\item \textbf{\texttt{softplus(z)}} (helper):
  $\log(1 + e^z)$, evaluated as $\max(z, 0) + \log(1 + e^{-|z|})$
  to avoid overflow. Used inside \texttt{GaussianNLLLoss} to map a
  real-valued network output to a positive scale parameter. Derivative
  is the sigmoid.
\item \textbf{\texttt{MSELoss}}: identity link, mean squared error.
  Single-output regression. The $\tfrac{1}{2}$ convention makes the
  gradient $z - y$, matching the canonical $\hat{p} - y$ pattern of
  the other losses.
\item \textbf{\texttt{PoissonNLLLoss}} (ten lines): single logit, log
  link. $L = e^z - yz$ (omitting $\log y!$).
  $\partial L / \partial z = e^z - y$. \texttt{probs} returns $e^z$.
\item \textbf{\texttt{GaussianNLLLoss}} (fifteen lines): two logits,
  $\mu = z_\mu$, $\sigma = \mathrm{softplus}(z_s)$.
  $L = \tfrac{(y-\mu)^2}{2\sigma^2} + \log\sigma$.
  Gradient derived in \Cref{sec:heteroscedastic}. \texttt{probs}
  returns $(\mu, \sigma)$.
\end{itemize}

Gradient-check residuals on a Tanh MLP (seed~42): MSE
$2.3 \times 10^{-10}$, Poisson $9.6 \times 10^{-11}$, Gaussian
$1.1 \times 10^{-10}$. All well below the $10^{-4}$ tolerance.

Not in the library: Beta NLL and the MDN. Both are natural extensions
(see \Cref{sec:link-likelihood,sec:mdn}), not missing infrastructure.
Beta NLL needs only \texttt{math.lgamma}; the MDN composes existing
primitives. They are the next things to build, not gaps.

\section{Handoff}\label{sec:handoff}

This chapter took up the output-head axis of inductive bias in full. The
unifying frame: every supervised network is a maximum-likelihood
estimator under an assumed output distribution. The canonical-link
theorem (\Cref{thm:canonical-link}): with the canonical link, the
gradient of the NLL is always $\hat{p} - y$. Three review pairings
from Chapter~1 (Gaussian, Bernoulli, Categorical), now named as members
of one family. Two new worked heads: Poisson for count data, and the
heteroscedastic Gaussian for calibrated, input-dependent uncertainty.
One example of link/likelihood independence (sigmoid plus Beta). A
pointer to mixture density networks. And the parallel-axes synthesis
showing how the architecture axis and the output-head axis compose.

The architecture axis continues in the next chapter, on convolutional
networks: the prior that nearby inputs matter together and the same
feature detector should fire anywhere. After CNNs, the following
chapters cover fixed-context language models, recurrent networks, and
the Transformer, each encoding a different structural prior. Every
architectural choice in those chapters composes freely with any of the
output heads in this chapter: a language model is a Transformer body
with a Categorical head over a vocabulary; pixel-wise depth estimation
is a CNN body with a Gaussian head; photon-counting imaging is a CNN
body with a Poisson head. The two axes multiply.

The book closes with a chapter on reinforcement learning, where the
training signal shifts from a per-example label to a scalar reward over
whole trajectories. The heads-as-bias frame still applies there (a
policy head is a softmax over actions; a value head is identity plus MSE
on returns), but the supervised-learning frame, the frame this chapter
closes, no longer holds. That shift is the genuinely different third
learning paradigm.
